# Assignment 05

**To submit this assignment, upload the full document to Quercus, including the original questions, your code, and the output. Submit your assignment as a knitted `.pdf`.**

Load these packages before you start:

```{r message=FALSE, warning=FALSE}
library(car)
library(ggfortify)
library(tidyverse)
```

**1. In this exercise, we will be analyzing floral trait data using a base R data set `iris`.** (4 marks)

```{r multivariates stats} 
head(iris)
```

This data set consists of 4 floral trait measurements (sepal length and width, and petal length and width) for 3 species of iris. 

a. Calculate the correlations between each variable and produce the pairs plot showing the relationships. Which pairs of variables show high collinearity? (0.75 mark)

b. Perform a principle components analysis on the floral traits in the `iris` data set (use a correlation matrix). Visualize your result in a biplot. (1 mark)

**NOTE: the `select` function in `dplyr` clashes with a function with the same name in the `raster` package. Be sure to specify to `R` which package you are calling the function from in your answer.**

c. Take a look at the trait loadings (`rotation`), and explain the loadings for PC1 and PC2 in plain English. (0.75 mark)

d. Colour the individual entities in your biplot by `species`. Explain the floral characteristics of the species "virginica" based on this plot. (1 mark)

e. Based off of the cumulative proportion of variance explained by each axis or the scree plot, which axis (axes) would you retain for downstream analyses and why? (0.5 mark)

**2. In this exercise, you will create and clone a _new_ GitHub repo onto your machine (call it assignment5_folder), use the command line to run an R script in this directory, modify the files that are returned by the R script, and push the results to remote repo.**

a. Go to _your_ GitHub (i.e., to github.com/[yourusername]). Make a new repo called assignment5_folder. (0.25 mark)

b. Clone your GitHub repo onto your machine. (0.25 mark)

c. Make an R script that uses a function to write a .csv file (named assignment5_file.csv) with two columns. The first column should be named x and should contain all numbers between 0 and 100. The second column should be named y and contain draws of a Normal random variable with mean 0 and _variance_ 0.001. In your call to `write.csv`, be sure to specify `row.names=FALSE`. (1 mark)

d. Use the command line to modify assignment5_file.csv so that the first entry in the x column is equal to 7 instead of 1. In your answer, specify what commands you ran (and in what order) to access and change the file. (1 mark)

e. Add, commit, and push the changes to the assignment5_folder repo to GitHub. As before, in your answer, specify what commands you ran (and in what order) to do this. (1 mark)

f. Provide one reason we use the command line, and one reason we use GitHub. (0.25 marks each)