# Assignment 03

For this assignment, we will be using the same `beaver1` dataset that we used in last week's assignment. Run the code below to create a categorical variable of the `activ` column, as we did for the last assignment. This will make dplyr recognize that there are only two levels of activity.

```{r eval=FALSE}
library(tidyverse)
beaver1_f <- beaver1 %>%
  mutate(factor_activ = factor(activ))
```

**1. Beaver body temperature (6 marks)**

a.  Create a histogram to visualize the distribution of the beavers' body temperatures, separating the temperature data based on the beaver's activity level. Choose an appropriate bin width for your histogram. (1 marks)

b.  Quantitatively describe the properties of the distribution (mean, standard deviation, median, range, and skewness). (0.5 marks)

c.  Suppose that the beaver body temperatures are normally distributed with mean $\mu$ and variance $\sigma^2$. Find the maximum likelihood estimates for $\mu,\sigma^2$ and report $90\%$ confidence regions for each parameter.[^1] Show your calculations (code) and explain your conclusions in plain English. Be sure to discuss how wide the confidence intervals are and why this may be the case. (2.5 mark)

d.  Perform $t$-tests to examine whether beaver's body temperature differ by activity level. Repeat this test after removing the outliers. Explain the results of your $t$-tests in plain English. Finally, discuss if we should trust the conclusions of the $t$-tests preformed: are the assumptions (normality of the data, constant variance between groups) violated? (2 marks)

[^1]: Recall that a confidence interval for a parameter can be constructed as follows: 1) evaluate the likelihood at the MLEs for all other parameters. This returns a slice of the likelihood function as a function of the parameter of interest. 2) Determine what values of the focal parameter $\theta$ are such that $\lambda_{LR} -2(\ln L(\theta)-\ln L(\hat{\theta}_{\text{MLE}})) < \chi^2_c$, where $\chi^2_c$ is cutoff for admissible values based on $100(1-\alpha)\%$-ile for a $\chi^2$ distribution with one degree of freedom. Hint: modify the code from class to do this.

**2. Explain the logic of likelihood (0.25 mark)**

a.  In your own words, explain what is a *parameter* of a statistical model, and explain what the method of maximum likelihood does in order to estimate that parameter.

**3. Outliers (1 mark)**

a.  In the beaver1 dataset, there are some particularly high/low body temperature measurements. How many data points laid out of the "normal" range, as defined by ggplot's boxplot? (0.5 marks)

b.  Give an example of a systematic and a random error that could have influenced these values. (0.5 marks)

**4. Hypothesis testing: some additional reading (0.25 mark)**

a.  Read [these notes](https://eeb313.github.io/tests-assignment-3.html) on some common statistical tests (chi square, permutation, etc.) used in EEB!

**5. Thanksgiving! (0.5 mark)**

\vspace{6pt}

Run and print this code chunk:

```{r, eval=FALSE}
install.packages("praise")
```

And then this:

```{r eval=FALSE}
library(praise)
replicate(100, praise())
```
