# Challenge assignment

```{r, echo=F, message=F, warning=F}
require(tidyverse)
theme_set(theme_bw())
```

**To submit this assignment, upload the full document to Quercus, including the original questions, your code, and the output. Submit your assignment as a knitted `.pdf`.**

Part of being an effective scientist involves being able to solve problems you have not encountered before. This is certainly true of programming as well, where problems are typically solved by furious bouts of Googling, reading documentation, and trial-and-error of proposed solutions. In this assignment, like previous ones, you will be evaluated on your ability to solve data manipulation and analysis tasks. However, unlike previous assignments, some of the solutions to the problems will require more research and effort on your part. It may require the use of packages and techniques not explored in class, but all problems are solvable, often with only a few lines of code. By now, you should all have the terminology required to work through and search for solutions to the problems below. Good luck, and let us know if you have any questions!

**1. Estimation of the population mutation rate [4.5 marks]**

As we discussed in class, the [Wright-Fisher model](https://stephens999.github.io/fiveMinuteStats/wright_fisher_model.html) describes how, in a diploid population of constant size $2N$, the number of genes of type $A_i$ ($i=1,2$) change due to random sampling of gametes (i.e., genetic drift) and mutation between alleles at the locus under consideration.^[Recall that the rules of the WF model are very simple. The population size is constant, generations are discrete and non-overlapping, and children choose their parents (regardless of type) at random. This means that the number of $A_1$ individuals in generation $n+1$ is Binomial($N$,$p_{n}$) where $p_n$ is the frequency of $A_1$s in the previous generation.] The model is widely used in population genetics, and can be used to make inferences about the population mutation rate $\mu = 4 N u$ from present-day allele frequency data.^[To make inference under the model tractable, it is actually an _approximation_ to the discrete WF model is used, in which it is assumed the $N$ is large and the per-individual mutation rate is small.] The population mutation rate measures the number of mutants that enter the population per generation, which means it has units of individuals per time. **In this problem, you will use maximum likelihood to make inferences about this parameter!** 

**1a.** Read in the data. The observations are frequencies of the $A_1$ allele in 52 replicate populations. For the purpose of this exercise, we will treat the allele frequencies as independent. This is reasonable if there is no gene flow between populations sampled; otherwise, we may need to re-think this assumption.

```{r}
WF_inference_data <- read.csv("WF_inference_data.csv", row.names = NULL)
head(WF_inference_data)
```

It is well-known that, when the WF model has been run out for a long enough time and with a large enough population size, the allele frequencies follow a _Dirichlet distribution_:

$$f(p|\mu) = \frac{\Gamma(2\mu)}{(\Gamma(\mu))^2} (p(1-p))^\mu,$$
\vspace{3pt}

where $\Gamma(\cdot)$ is the Gamma function (a generalization of the factorial). One can evaluate this probability density at a _particular_ set of allele frequencies and value of $\mu$ by calling the following function:

```{r}
eval_dirichlet <- function(p, mu){ 
  # p = vector of allele frequencies, mu = single value of mutation rate
  return(
    (gamma(2*mu)*(p^mu)*((1-p)^mu))/(gamma(mu)^2)
  )
  # returns probability density of each allele frequency observation in vector p
  # for the value of mu specified
}

eval_dirichlet(c(0.81,0.9),0.0001) 
# for example, this is the value of the density at p=0.81,0.9 and mu = 0.0001
```

Importantly, and as in many other applications, the data are non-normal and we must make parametric assumptions that reasonably describe the processes by which the data may have been generated and sampled. Because theory predicts the allele frequencies are Dirichlet distributed for large enough populations and long enough times, we will make this assumption and proceed with maximum likelihood. Because the data are independent and all follow this distribution, the log-likelihood function is

$$\sum_{i=1}^{52} \ln f(p_i|\mu),$$
\vspace{3pt}

where $p_i$ is the allele frequency in the $i$th population and $\mu$ is a specific value of the mutation rate. The `eval_dirichlet` function evaluates $f(p|\mu)$ for allele frequencies that are supplied as a vector in its first arguement and a specific value of the mutation rate which is specified in its second.

**1b.** Using `eval_dirichlet`, write a function that calculates the log-likelihood of the allele frequency data from the 52 populations. The function should have an argument corresponding to the value of the population mutation rate $\mu$, which is called `mu` in the function which evaluates the Dirichlet density. [1 marks]

**1c.** Using the function from 1b, evaluate the log-likelihood across a range of values for the parameter. A suitable range to use is from 10 to 30. Plot the likelihood as a function of $\mu$ over this range. What is the maximum likelihood estimator for $\mu$? [2 marks]

**1d.** Build a $95\%$ confidence interval for $\mu$ by determining what values of $\mu$ are such that the log-likelihood is within 1.92 units of its maximum value, i.e., the value assumed when the estimator for $\mu$ is equal to its MLE. (The 1.92 quantity is the cutoff based on the approximate $\chi^2$ distribution of the likelihood ratio test statistic, which we have previously used to build confidence intervals and perform hypothesis tests.) [1 mark]

**1e.** Based on the confidence interval constructed in the previous question, test the hypotheses $H_0 \colon \mu = 10$ vs $H_1 \colon \mu \neq 10$ at significance level $\alpha = 0.05$. [0.5 marks]

**2. A mathematical model of the evolution of sex [5 marks]**

Approximately $99.9\%$$ of all of eukaryotes have the ability to reproduce sexually (i.e., are facultatively sexual). Why this is the case remains an important question in evolutionary genetics. **Here we will use a mathematical model of one hypothesis for the ubiquity of sex across eukaryotes: differences in extinction rates between sexual and asexual linages.**

Let $S(t)$ be the number of sexual species and $A(t)$ be the number of asexual species at time $t$. Both types of species spectate at per-capita rate $b$, sexual species go extinct at rate $d$, and asexual species go extinct at an _elevated_ rate $d+\delta$. Sexual species give rise to asexual species a fraction $p$ of the time, but asexual species never give rise to sexual species. Time is measured in thousands of years.

Under these assumptions, equations we could use to model the number of sexual and asexual species are

$$\frac{\text{d} S}{\text{d} t} = b (1-p) S - d S.$$
$$\frac{\text{d} A}{\text{d} t} = b A + b p S - (d+\delta) A$$

**2a.** Starting from ten sexual and ten asexual species, solve the model equations out to time $t=100$ and plot the fraction of species that are sexual through time. Use the following parameter values: $b=1, d = 0.7, \delta = 0.5, p = 0.1.$ What fraction of species are sexual in the long run given these parameters? [2 marks]

**2b.** Suppose $p=0.01$ and $b = d = 1$. Write a function to numerically solve the model equations for a specific value of $\delta$ and return the fraction of asexual species. Then loop over values of $\delta$ to determine the value of $\delta$ so that the long-run fraction of sexual species is $=0.999$. In other words, determine the value of $\delta$ so that the fraction of sexual species is equal to what has been estimated empirically. [1.5 marks]

**2c.** Explain the findings of 2b. For the chosen parameters, how much larger does the extinction rate of asexual lineages have to be to explain the fact $99.9\%$ of all eukaryotes are able to reproduce sexually? [1 mark]

**2d.** What do you think of the assumption some of speciation events in sexual species result in the formation of asexual species? Does it make sense, given the biological question at hand? Explain why or why not. [0.5 marks]

**3. Take a break [1 mark]**

Execute the follow code chunk:

```{r, eval=F}
library(praise)
replicate(100, praise())
```

**4. GLMs: power analysis and fitting [9.5 marks]**

The horseshoe crab _Limulus polyphemus_ has two male reproductive morphs; the smaller males have a a special appendage known, which are used to attach to female crabs. When female crabs dig a nest and lay eggs on the beach, the attached male can then fertilize the eggs. Alternatively, "satellite" males can crowd around nesting pairs and then fertilize eggs laid. In this question, you will

- do a _power analysis_ to determine the effect size, under a Poisson regression model^[This is a special kind of generalized linear model! The Poisson distribution models count data, like the number of satellite males at a beach.], is determined to be significant at least 99% of the time

- fit a Poisson regression model to satellite crab count data to determine if female traits are associated with more satellite crabs

- interpret the Poisson regression coefficients (be sure to make note of the units!)

- perform model selection!

The data are retrieved below. Variables include female color, spine condition, carapace width (cm), mass (kg), and number of satellite males.

```{r}
satellites_data <- read.csv("satellites.csv", row.names = NULL)
head(satellites_data)
```

**Before** we analyze the data, we will determine the effect size we can reliably estimate from the data at hand at a given significance level. This is typically done before collecting data, but for the purposes of demonstration we will have you do it right before diving into fitting a GLM to the data.

**4a.** Define statistical power in the context of regression (i.e., where the null hypothesis is that an effect size is zero and the alternative hypothesis is that it is non-zero) in your own words. [0.5 marks]

**4b.** The following function draws observations from a Poisson regression model $Y \sim \text{Poisson}(\lambda)$, where $\log \lambda = \beta_1 x$. The parameter $\lambda$ is the expected number of satellite males, and $\beta_1$ is the size of the effect of increasing female carapace width $x$ on $\log \lambda$. The function has one _scalar_ argument, `beta`. (This means `beta` = 3 is a reasonable input but `beta` = c(3,4) is not.) The function returns a data frame of trait values `x` in the first column and associated number of satellite crabs `y` in the second.

```{r}
n <- 173

data_generator <- function(beta){
  x <- sample(size = n, seq(20, 35, by = 0.001), replace = T)
  obs <- c()
  
  for (i in 1:n){
    obs[i] <- rpois(1, exp(beta*x[i]))
  }
  
  return(data.frame(x, y = obs))
}
```

Use this function to simulate draws from the statistical model when $\beta=0.0001,0.001,0.01,0.1$. Bind these observations into a data frame with a column that specifies the effect size associated to a given observation. Display the head of this data frame, and make a scatter plot of $y$ against $x$, i.e., the number of satellite crabs against carapace width, _faceted by effect size_. Allow the $y$ axis on each plot to be free. [2 marks]

**4c.** Write a function to determine, for a given effect size, the power of a Poisson regression of number of satellite males on carapace width. To do this, you will have to

- write a for loop to simulate `sims` = 1000 draws of the Poisson regression model via `data_generator()`
- fit a Poisson regression of number of satellite males on carapace width using each simulated dataset
- extract the $p$-values associated to each model.
- calculate outside of the for loop what proportion of the $p$-values which are $< 0.05$

The proportion of $p$ values which are $< 0.05$ is your estimate of power at the effect size, sample size, and significance level specified. This is because it gives, for a large number of draws from the Poisson regression model at a given effect and for $x$ distributed according to our expectations for carapace width, the fraction of fitted models for which the effect was determined to be $\neq 0$. [1.5 marks]

**4d.** Using the function from the previous question, estimate the power of the Poisson regression model for $\beta = 0,0.025,0.05,0.075,0.1$. Display the effect sizes and associated power estimates. At what effect size is the power for a Poisson regression with $n = 173$ observations >99%? [1 mark]

**4e.** Using the satellites data, fit a Poisson regression of the number of satellite males on carapace width. Report the intercept and slope coefficients, associated $p$ values, and AIC. [1 mark]

**4f.** Recall the description of the Poisson regression model in question 4b. How should we interpret the regression coefficient associated to carapace width? [1 mark]

**4g.** Plot the data and the fitted model from 4e. The regression line should be curvilinear on the data scale. [1 mark]

**4h.** Using the satellites data, fit the following Poisson regressions:

- the number of satellite males on female mass
- the number of satellite males on carapace width and female mass
- the number of satellite males on carapace width, female mass, and their interaction

Perform model selection: report which of these models (including the one from 4e) is the best fit to the data. **Report the model that is the best fit to the data and its AIC. Are there any models with AICs that are close? What does this mean?** [2 marks]