# Multivariate statistics

```{r message=FALSE, warning=FALSE}
library(car) # VIF 
library(tidyverse)
library(ggfortify) # fancy biplots
library(wesanderson) # fancy colours
library(vegan) #  tools for diversity analysis, ordination, calculating dissimilarity and descriptive community ecology
```

## Lesson intro

Most things in the natural world are inter-related and correlated, where many
trends and observations tend to change in a predictable way together. This is
personally one of the things that I find exciting about biology--the sheer
complexity of interactions in natural systems.

But this also makes our lives as scientists challenging, because we then have to
distill, what is sometimes called a "causal soup", into variables of interest
and then figure out a meaningful analysis that can tease apart which effects
came from where.

And this is where it gets complicated---how do we accurately represent the true
complex and multidimensional nature of ecological systems without breaking our
brains?

Cue multivariate statistics.

We use multivariate techniques to reduce complex situations into more manageable
ones and to deal with the fact that patterns are correlated with other patterns,
which are in turn driven by a number of interacting ecological processes.

Up until now we've been working with univariate methods -- when there is a
single response variable is of interest, or when we have multiple predictors but
we are mostly interested in their independent effects on the response.

In contrast, multivariate statistics is a family of statistical tools used to to
handle situations where the predictor and/or the response variables come in
sets, allowing us to distill through some of that correlated (and sometimes
redundant) information and to analyze these variables simultaneously, as they
should be.

Unfortunately (?), it is not within the scope of this class (or even this
course!) to teach all of this. Multivariate statistics area large discipline in
their own right. They are a flexible tool with a lot of variations - which can
make it challenging to choose the correct technique for an analysis.

Today, we will focus on a couple of the most common techniques that you are
likely to come across in the papers you read and/or use in your own work. This
is a quick overview and we necessarily have to have to skip over some of the
details---if want to know more or have a specific question related to data, we
are happy provide readings or to chat during office hours.

### Lesson outline

-   Examples of multivariate statistics using well-behaved data

    -   PCA and PCoA

    -   Discuss caveats and analytically choices embedded in these methods

-   Work through an analysis of real-world data

    -   Assess multicollinearity

    -   PCA

### Learning objectives

-   Learn conceptual theory behind common multivariate techniques

-   Learn how to run and interpret the output from PCA and PCoA

-   Learn how to assess multicollinearity

-   Be able to apply the above techniques on real-world data

-   Your types of data and biological questions drive your types of statistical
    analysis

-   **Recognize that the choices required in a statistical pipeline can
    drastically affect your results, but there is no one right way of analyzing
    your data**

### A bit of theory[^1]

[^1]: This explanation was largely lifted verbatim from Borcard, D., Gillet, F.
    and Legendre, P. 2011. *Numerical Ecology with R.* Chapter 5, page 116.

Multivariate statistics roughly split into unconstrained vs constrained
ordination techniques.

Unconstrained ordinations have no a priori groupings or relationships in the
variables or observations. Common examples include: PCA, PCoA, cluster analysis,
correspondence analysis (CA), factor analysis and NMDS. We will focus on this
broad category today.

Constrained ordinations specify a priori groupings. Examples include
discriminant analysis, canonical variate analysis, canonical correlation
analysis, and redundancy analysis.

Unconstrained ordinations generally work by the extraction of the eigenvectors
of an association matrix. For example, imagine your data containing n objects
and p variables. The n objects can be represented as a cluster of points in the
p-dimensional space. This cluster is generally not spheroid: it is elongated in
some directions and flattened in others. These directions are not necessarily
aligned with a single dimension (= a single variable) of the multidimensional
space.

The direction where the cluster is most elongated corresponds to the direction
of largest variance of the cluster and this is the first axis that an ordination
will extract. The next axis to be extracted is the second most important in
variance, provided that it is orthogonal (linearly independent, uncorrelated) to
the first one.

The process continues until all axes have been computed. When there are a few
major structures in the data (gradients or groups) and the method has been
efficient at extracting them, then the few first axes contain most of the useful
information, i.e. they have extracted most of the variance of the data. In that
case, the distances among sites in the projection in reduced space are
relatively similar to the distances among objects in the multidimensional space.

How the methods do this specifically varies and depends on the types of data and
specific methods you are using.

For more details see
[this](https://link.springer.com/content/pdf/10.1007/978-1-4419-7976-6.pdf)
chapter in Borcard, D., Gillet, F. and Legendre, P. 2011. *Numerical Ecology
with R.*

## Let's practice: toy example

For this section we will be using soil chemistry data from the `vegan` library.

```{r varechem data}
data(varechem)
head(varechem)
str(varechem)
```

This data frame is composed of 24 rows that correspond to different sites and 14
columns, which correspond to the different chemicals measured at each site.

### A note on distance matrices

Many multivariate methods rely on being able to represent your observations as
being more or less similar to each other (similarities or distances). This
makes, because if you're trying to group observations by variance along an axis,
you need to know how far apart each observation is from other observations.

This is easy to do (and visualize) with quantitative data---a distance is just
the numeric difference between the observations along an axis. To deal with
negative numbers and multiple dimensions, rather than subtracting one value the
other, we instead compute **euclidean distances**.

For visualization, let's compute the euclidean distances between the first four
soil chemicals. We will use the `vegdist` function in `vegan`*.* `vegdist` is
handy because it allows us to specify which type of distance matrix we want to
use.

```{r euclidean distance}
?vegdist
# let's just look at the first column
euclidean.distance <- vegdist(varechem[, 1:4], method="euclidean") 
# lets see it 
euclidean.distance
```

Note that the output here is an object of class "distance" that shows all the
pairwise euclidean distances between all cells in the data.

Thinking about distances are more challenging to do if you have some data that
is categorical (what is the similarity along an axis of apples to oranges?).
Luckily there are different types of distance matrices that can handle different
types of data.

Another really common type of distance matrix is the Gower distance, which can
handle categorical and mixed data, and so is very useful. For more detail on
distance matrices read
[this](https://chrischizinski.github.io/snr_r_notebooks/distancematrices/).

Okay, now we're ready to do some statistics!!

### PCA

One of the most common (unconstrained) ordination techniques used in EEB is the
*Principle components analysis* (PCA). A PCA seeks to find a new coordinate
system (PC axes) that best accounts for the variation in a multivariate data
space, thus reducing dimensionality in the process but still accounting for as
much of the original variation as possible. Critically, a PCA assumes a linear
relationship between your variables, can only take quantitative data, and
requires more rows than columns in the data itself.

The procedure to conduct a PCA is as follows:

1.  Imagine a 14-dimensional object with each of our soil measurements as an
    axis (k=14).

2.  Plot our sites (n=24) in this 14-D space, and then find the Euclidean
    distance between those points (do not despair, `R` will do the actual
    calculations for us).

Let's visualize this (but in less dimensions because we are human). One way to
do this is with a `pairs` plot in base `R`, which plots the relationships
between all variables (columns) as a scatter plot.

```{r visualize the correlations}
pairs(varechem[, 1:4]) #only the first 4 columns for simplicity
```

*Which pairs of chemicals are tightly related to each other? Which are not
closely related?*

3.  After you have the distances between points, we can use procedures that are
    similar (in concept) to least squares to find a "best-fit line" through our
    data points in multidimensional space (again, `R` will take care of all
    this). This will be our first principle component (PC) axis and it explains
    the most variance in our data.

4.  Find another best-fit line through the data under the condition that it has
    to be orthogonal to the first axes (i.e., at a 90° angle, meaning that these
    two axes are as unrelated to each other as possible). This will be our
    second PC axis. [See here for visual PCA axis.](https://i.stack.imgur.com/JZF5m.gif)

5.  This process continues until we have k PC axes.

Okay, let's work through some of the details in an example.

We will use the function `prcomp`. At its most basic, `prcomp` takes your data
and then you specify if you want to `scale = TRUE/FALSE`. The scale command
specifies whether you want to run the pca on a correlation matrix or a
covariance matrix. This sounds complex but it boils down to a simple idea.
Choosing `scale = TRUE` means you are selecting a correlation matrix which
standardizes your data to have unit variance ie puts all your variables on the
same scale. Choose this option if your data contains multiple types of units.

```{r run the pca}
varechem.pca <- prcomp(varechem, scale. = TRUE)
```

Phew, that wasn't too bad. Let's take a look at our varechem.pca object

```{r summary}
summary(varechem.pca) # this is ugly becuase of the linebreaks, look at in console, it's easier
```

This gives us a table *summary* of variance explained by each axes. `R` even did
the math for us and showed cumulative variance explained as we move to the
higher dimensions, so convenient! We see that the first two PC axes already
explained close to 60% of the variation of the data. But it doesn't give us any
information about our variables or sites!

Let's have a look at this object in another way.

```{r structure}
str(varechem.pca)
```

We see that this object is structured as a five-item list. We mostly care about
two of these items:

-   Loadings -- appropriately named rotation.

-   Site scores -- lazily named x.

To extract items from a list, use the \$ notation.

```{r loadings varechem}
varechem.pca$x # this gives us the particular location (ie the coordinates) of each of our sites in the multidimensional space formed by the PC axes

varechem.pca$rotation # this tells us which of our soil chemicals are contributing most strongly to each PC axis (ie which variables are being summarized by each newly created PC axis)

```

We can interpret these outputs directly, specifically the rotations.

```{r interpret loadings}
varechem.pca$rotation[ ,1:4] # let's just look examine the first 4 PC axes
```

The way we interpret this is by looking, within each PC axis, for the variables
that have the largest absolute values. These are the variables that are
contributing variation to this axis.

For example, PC1 sets up an axis of increasing concentrations of P, K, Ca, Mg,
S, Zn. PC2 is driven largely by Al, Baresoil, Humdepth and pH. Note that the
positive and negative numbers indicate a positive and negative relationship (eg
along PC1 P and Fe are negatively correlated, sites that have high P tend to
have low Fe and vice versa).

This is much, much easier to see visually. We can do this with the base `R`
function `biplot`, which is a bit ugly.

```{r biplot}
biplot(varechem.pca)
```

On this figure, there are 2 sets of information. Each site is plotted (these are
the numbers). As well, the variable loadings are represented as arrows and tell
us how each variable is contributing to each axis. The clustering of sites along
the axes (and rotations) tell us which sites are similar in terms of their
chemical composition. Eg, we can see that sites 19, 20, 27 and 28 all have a
similar chemical composition. We can also see clearly that along PC2 sites with
low pH also tend to have low Al, while they have high Humdepth, Baresoil and Mn.
Future analyses might relate the soil characteristics of these sites to other
environmental data (history of land use, pollution etc)

We can do this with the function `autoplot` in the `ggfortify` package, which
takes advantage of ggplot style plotting.

```{r autoplot}
autoplot(varechem.pca) +
  theme_classic()
```

The simplest version of `autoplot` displays less information than our `biplot`
function. On this figure we can see that we are plotting PC1 by PC2 and each
point on the figure is each of the rows aka sites in our data. `autoplot`
handily also shows us the proportion of variance explained by each of these PC
axes. Now lets add the loadings.

```{r add the loadings }
# Remember to tell R where the original data is coming from! 
# Also, `autoplot` will refuse to understand you if you spell "colour" the non-Canadian way
autoplot(varechem.pca, loadings=TRUE, loadings.label=TRUE, 
         loadings.colour="grey30", 
         data=varechem) +
  theme_classic()
```

But do we make biplot for each of our PC axes??? Nope, this would take too long
and each subsequent axis explains less of the variation in our data... ie it
becomes less important. As well, we often perform PCA to simplify our data from
many variables to few.

So the next step is to pick how many axes show meaningful information.

We can do this with a scree plot. This shows us the amount of variation
explained by each axis. As we know, each axis explains subsequently less
variation. We use scree plots to help us determine at which point the axes stop
adding much new information. We can do this visually or quantitatively. We will
just go over the visual method as determining which axes to keep is pretty
subjective and depends on your system and question of interest.

```{r scree plot}
plot(varechem.pca)
```

Each bar is a PC axis, starting from PC 1. In a scree plot, what you want to
look for is a big jump between subsequent axes. Its subjective, but based on
this scree plot I would probably be inclined to keep the first 2 or 3 PC axes.
As well, we can look at the cumulative proportion of explained variance:

```{r}
summary(varechem.pca)$importance[,1:6]

```

And we can see that by PC3 we have explained \~72% of the variation in the data,
which is not bad (not great either, but not bad). By the time we get to PC4,
each individual axis is explaining less than 10% of variation in the data, so
adding them, and conversely excluding them, isn't going to have a large effect.

\*Note 1: there are many functions across many libraries for running a pca:
`prcomp`, `princomp`, `rda`, `dudi.rda`. Each method makes a slightly different
choice about how the underlying math is performed. `prcomp` and `rda` are both
very common choices.

\*Note 2: for those of you working with phylogenetic data: you may want to
perform a phylogenetic PCA (pPCA). The function `phyl.pca` is available in
package `phytools`.

For further readings see:

-   [PCA: a visual
    explanation](https://towardsdatascience.com/principal-component-analysis-pca-explained-visually-with-zero-math-1cbf392b9e7d)

-   [PCA explained using a metaphor with wine, dinner, grandmothers, and
    animations](https://stats.stackexchange.com/questions/2691/making-sense-of-principal-component-analysis-eigenvectors-eigenvalues)PCoA

### PCoA

A PCoA (principal coordinates analysis) is a more flexible form of a PCA that
can take any type of distance matrix. Remember that distance matrices allow you
to express differences between your data points so by specifying the type of
distance matrix, so this means that PCoA can handle a mix of data types---which
is very handy if you have both categorical and numeric data. Note that your
choice of distance matrix can have a large effect on the conclusions you draw in
your analysis--make sure to read the literature and learn what best practices
there are for you specific data types.

You interpret the output of a PCoA the same way as a PCA. Common functions are
`cmdscale` in base `R`, `wcmdscale` in `vegan`, and `pcoa` in `ape`.

In sum, both PCA and PCoA preserve the distances between your data points--for
this reason they are commonly called "rotations"--these methods do not change
the underlying relationships in your data, they just group variation in data
across many dimensions and then make a nice tidy output that our brain's can
interpret.

## Let's practice: real life example

For this section of the class, we will try to apply what we've learned above on
a real world data set. We'll be walking through the analysis and the choices we
make and then at each section, I will give you a minute or so to try and either
a. shout out the answer or b. write the code.

We are working with a cool dataset that was collected by Vicki as part of her
PhD work outside of Churchill, Manitoba. Vicki generally studies invasive
species in the tundra and the data we are working with today is part of a larger
project that investigates the effect of soil on plant invasions.

The key question we are trying to answer today is if soil chemicals, which we
expect to vary across a landscape according to human presence and land use, can
predict whether or not invasive species are present.

Let's read in our data and take a look at it!

```{r Read in data}
soil.data <- read_csv("soil.data.csv")
head(soil.data) 
str(soil.data)

```

Okay, so in this data, we have columns that contain our environmental data: the
concentration of various chemicals in the soil, if there are invasive species
(categorical, two levels, invaded/uninvaded), and then a bunch of spatial
information that relates each site to a broader geographic region ( 3 levels,
town, airport, CNSC). We may expect that each of these broader regions will have
a particular soil chemical composition, which then should (according to our
question) affect the presence of invasive species.

More formally then, we are testing for a relationship between location, P, C, N,
and avg_pH (our explanatory variables) and invaded_status (our response
variable).

Before we get started, let's do some data organization.

```{r organize the data}
names(soil.data)

# convert response into 1s and 0s
soil.data$binary.invaded_status <- ifelse(soil.data$invaded_status=="invaded", 1, 0)

# make sure that location is a factor
soil.data$location <- as.factor(soil.data$location)

```

### Multicollinearity

The first thing we need to do is assess the relationships between our
explanatory variables (location, P, C, N, avg_pH) and check for
**multicollinearity**.

Multicollinearity refers to a situation in which **two or more** explanatory
variables are highly linearly related (See
[here](http://web.vu.lt/mif/a.buteikis/wp-content/uploads/PE_Book/4-5-Multiple-collinearity.html)
for a mathematical explanation). This can wreck havoc in an analysis because
collinear predictors generate unstable and unreliable regression coefficients
that have large standard errors.

So the first step in this analysis is to visualize all our environmental data
and assess whether any variables are highly correlated with each other (ie are
collinear).

#### Correlations

There are a couple of ways of assessing collinearity of our predictor variables.
One common way you may have heard of is to calculate the correlation between the
variables. We do this using `cor` and can visualize the relationships using
`pairs`. Note that this method only works for numeric data (ie it can't handle
categorical variables).

```{r visualize data }
pairs(soil.data[,7:10])
cor(soil.data[,7:10])

```

We can see from this that C and N are highly related. When we look at the
correlations, this linked relationship is confirmed. As a rule of thumb,
ecologists generally report an r\>=0.70 as being collinear. Note that this
threshold is a bit arbitrary and different sub-disciplines have different
traditions.

Based off this output, we could choose to drop either C or N from the analysis
as they will explain similar variation in the data. Generally, we would choose
to drop whichever variable we were less interested in scientifically.

#### VIF = Variance Inflation Factor

The other common way of assessing collinearity is with the VIF. This method can
handle both categorical and numeric data.

The VIF is calculated for each predictor by doing a linear regression of that
predictor on all the other predictors, and then obtaining the *R^2^* from that
regression. The VIF is just 1/(1-*R^2^*). This is why it's called the variance
inflation factor - it estimates how much the variance of a coefficient is
"inflated" due to linear dependence with other predictors.

VIF's start at 1 and have no upper limit. In general:

-   VIF = 1 means no correlation between predictor and any other predictor
    variables in model

-   1 \< VIF \< 5 suggests moderate collinearity

-   VIF \> 5 means severe collinearity

Note that there are some cases in which high VIF values can be ignored, for
example for polynomial terms (see
[here](https://statisticalhorizons.com/multicollinearity/) for a detailed
discussion).

VIF is used on model rather than on the raw predictors. So let's fit our model
and include location (town vs airport) and see if any of our predictors are
collinear. A VIF above 3 (or also commonly 5) indicate high collinearity.

Note that for 28 observations this model is also very over fit (too many
predictor terms to get general model parameters), but let's ignore that for now.

```{r fit model and check VIF}
model_collinear <- glm(binary.invaded_status ~ location + P + C + N + avg_pH, 
                      data=soil.data, 
                      family="binomial")
car::vif(model_collinear)

```

We have multiple levels in our categorical predictor (airport/town/CNSC) and so
we are returning GVIF which are produce better estimates when the degrees of
freedom \> 1. We interpret GVIF the same was as VIF and apply the same rule of
thumb.

*Which variables are collinear? Does this match our conclusion from using cor()
above?*

Just as above, we can see the C and N are collinear and have a high VIF.
Generally, best practice is to drop the term with the highest VIF and rerun, and
keep going, dropping terms until all VIF are below the threshold for
collinearity.

```{r fit model and check VIF again}
model_collinear_dropN <- glm(binary.invaded_status ~ location + P + C + avg_pH, 
                      data=soil.data, 
                      family="binomial")
vif(model_collinear_dropN) 
```

*How does VIF change if drop N? How does the output change? What can we conclude
from this?*

Note that dropping variables due to collinearity isn't p-hacking. In your
methods, you would report all the predictors and your method of checking for
collinearity, and what threshold you use, and which variables you exclude.

Okay, so what happens though if N is actually our variable of interest and we've
just gotten rid of it from the model??? Do we panic??

No! Luckily we've just learned about multivariate techniques for grouping
correlated data into common axes of variation.

#### PCA

*Which method might we use on this data?*

PCA is also useful in this case for exploring how these different soil
measurements are organized across sites.

*Should we use a covariance or correlation matrix?*

```{r pca on soil data }
# pca on data
# each row is a site
names(soil.data)

# make sure to select columns that you want to include in the PCA!
pca.soil <- prcomp(soil.data[, c(7:10)], scale. = TRUE) # use corr bc multiple units
```

Now let's look at the output.

```{r soil loadings and scree plot}
# look at importance of components
plot(pca.soil)

```

*How many axes should we keep?*

```{r loadings soil}
# let's look at loadings
pca.soil$rotation
```

*So how would we interpret this? Let's start with PC1.*

Now let's visualize, it'll be much easier to interpret! We're just going to use
`autoplot` here, but remember the base plot function `biplot` also works.

```{r soil biplot }
autoplot(pca.soil, loadings=TRUE, loadings.label=TRUE, 
         loadings.colour = "grey30",  # changes the colour of the arrows
         data = soil.data, 
         colour="location", # colour obs by location
         loadings.label.hjust = 1.2) + # moves loading labels so easier to read
  theme_classic() +  # make fancier (removes box)
  scale_color_manual(values= wes_palette("FantasticFox1", n = 3))  # pretty colours!
```

*What jumps out to you about this figure? What are some of the things this tells
us?*

1.  PC1 explains most of variation and is doing most of heavy lifting
2.  P, C, and N are all highly correlated together (arrows of similar length and
    pointing in same direction)
3.  pH is relatively uncorrelated (orthogonal to P, C, N)
4.  P, C, N load strongly on PC1, setting up an axis of more to less
    concentration
5.  pH loads mostly (but not exclusively) on PC2
6.  If you squint *maybe* the different locations are grouped in PC space... but
    more likely it looks like location doesn't explain where the data (sites)
    fall out in relation to P, C, N, and pH (ah well, this is what happens when
    we work with real data!)

Let's run our same logistic model again to test if soil characteristics explain
the relationship between if a site and invasive species or not. This time,
instead of using all the soil data, we will use PC axes.

But first, some (more) data organization.

```{r}
# see each observation and it's location along PC axes
pca.soil$x

# bind the observations to the data, note that `R` keeps everything in the same order, so this is easy
soil.data$PC1 <- pca.soil$x[ ,1]  # pull PC1
soil.data$PC2 <- pca.soil$x[ ,2]  # pull PC2
head(soil.data)
```

#### Modelling

Okay, now let's run the model.

```{r run logistic model }
model_PCA <- glm(binary.invaded_status ~ location + PC1, 
                      data = soil.data, 
                      family = "binomial")
vif(model_PCA) # nice, this is what we want to see

```

```{r interpret output }
summary(model_PCA)
```

Ah well, it looks like there's no relationship between soil characteristics and
invasion.

We can actually see this very clearly if we just plot our data.

```{r final plot }
ggplot(soil.data, aes(PC1, binary.invaded_status, colour=location)) +
  geom_point(size=3, alpha=.8) +
  theme_classic() +  
  scale_color_manual(values= wes_palette("FantasticFox1", n = 3))  +
  labs(y="Invasion success")
```