# Introduction to R: assignment, vectors, functions, strings, loops

*Note: This lecture content was originally created by voluntary contributions to
[Data Carpentry](https://datacarpentry.org) and has been modified to align with
the aims of EEB313. Data Carpentry is an organization focused on data literacy,
with the objective of teaching skills to researchers to enable them to retrieve,
view, manipulate, analyze, and store their and other’s data in an open and
reproducible way.*

The above paragraph is made explicit since it is one of the core features of
working with an open language like R. Many smart people willingly and actively
share their material publicly, so that others can modify and build off of the
material themselves.

By being open, we can "stand on the shoulders of giants" and continue to
contribute for others to then stand on our shoulders. Not only does this help
get work done, but it also adds to a feeling of community. In fact, there is a
common saying in the open source world:

> I came for the language and stayed for the community.

This saying captures the spirit, generosity, and fun involved in being a
part of these open source projects.

-----

## Lesson Preamble

> ### Learning Objectives
>
> - Define the following terms as they relate to R: call, function, arguments, options.
> - Use comments within code blocks.
> - Do simple arithmetic operations in R using values and objects.
> - Call functions and use arguments to change their default options.
> - Define our own functions
> - Inspect the content of vectors and manipulate their content.
> - Create for-loops
> - Describe what a data frame is.
> - Load external data from a .csv file into a data frame in R.
>
> ### Lecture outline
>
> - Setting up your R Notebook (10 min)
> - Creating objects/variables in R (10 min)
> - Using and writing functions (15 min)
> - Vectors and data types (10 min)
> - Subsetting vectors (15 min)
> - Missing data (10 min)
> - Loops and vectorization (10 min)
> - Data set background (10 min)
> - What are data frames? (10 min)

-----

## Setting up the R Notebook

Let's remove the template RStudio gives us, and add a title of our own.

```yaml
---
title: Introduction to R
---
```

This header block is called the YAML header. This is where we specify
whether we want to convert this file to a HTML or PDF file. This will be
discussed in more detail in another class. For now, we just care about
including the lecture title here. If you are interested in playing with
other YAML options, check out [this guide](https://bookdown.org/yihui/rmarkdown/html-document.html).

Under this sentence, we will insert our first code chunk. Remember that you
insert a code chunk by either clicking the "Insert" button or pressing
<kbd>Ctrl/Cmd</kbd> + <kbd>Alt</kbd> + <kbd>i</kbd> simultaneously. To run
a code chunk, you press the green arrow, or <kbd>Ctrl/Cmd</kbd> +
<kbd>Shift</kbd> + <kbd>Enter</kbd>.

## Creating objects in R

As we saw in our first class, you can get output from R simply by typing math in
the console:

```{r}
3 + 5
12 / 7
```

However, to do useful and interesting things, we need to assign _values_ to _objects_.

```{r}
x <- 3
x + 5
```

You can name an object in R almost anything you want:

```{r}
joel <- 3
joel + 5
```

#### Challenge

So far, we have created two variables, `joel` and `x`. What is the sum of these variables?

```{r, include=FALSE}
joel + x
```

### Some tips on naming objects

- Objects can be given any name: `x`, `current_temperature`, `thing`, or `subject_id`.
- You want your object names to be explicit and not too long.
- Object names cannot start with a number: `x2` is valid, but `2x` is not. 
- R is also case sensitive: `joel` is different from `Joel`.
- Avoid using the names of existing functions (e.g. `mean`, `df`). You can check whether the name is already in use by using tab completion
- Generally good to use underscores (`_`) to separate words in variable and function names

It is also recommended to use nouns for variable names, and verbs for function names. It's
important to be consistent in the styling of your code (where you put spaces,
how you name variables, etc.). Using a consistent coding style[^coding_style]
makes your code clearer to read for your future self and your collaborators.
RStudio will format code for you if you highlight a section of code and press
<kbd>Ctrl/Cmd</kbd> + <kbd>Shift</kbd> + <kbd>a</kbd>.

[^coding_style]: Refer to the [tidy style guide](https://style.tidyverse.org/index.html) for which style to adhere to.

### Preforming calculations

When assigning a value to an object, R does not print anything. You can force R
to print the value by using parentheses or by typing the object name:

```{r}
weight_kg <- 55    # doesn't print anything
(weight_kg <- 55)  # but putting parentheses around the call prints the value of `weight_kg`
weight_kg          # and so does typing the name of the object
```

The variable `weight_kg` is stored in the computer's memory where R can access
it, and we can start doing arithmetic with it efficiently. For instance, we may
want to convert this weight into pounds (weight in pounds is 2.2 times the
weight in kg):

```{r}
2.2 * weight_kg
```

We can also change a variable's value by assigning it a new one:

```{r}
weight_kg <- 57.5
2.2 * weight_kg
```

This means that assigning a value to one variable does not change the values of
other variables.  For example, let's store the animal's weight in pounds in a
new variable, `weight_lb`:

```{r}
weight_lb <- 2.2 * weight_kg # Actually, 1 kg = 2.204623 lbs
```

and then change `weight_kg` to 100.

```{r}
weight_kg <- 100
```

#### Challenge 

What do you think is the current content of the object `weight_lb`? 126.5 or 220?

```{r, eval=FALSE}
weight_lb
```

#### Challenge

What are the values after each statement in the following?

```{r, eval=FALSE}
mass <- 47.5
age  <- 122
mass <- mass * 2.0      # mass?
age  <- age - 20        # age?
mass_index <- mass/age  # mass_index?
```

## Functions and their arguments

### Understanding functions

Functions can be thought of as recipes. You give a few ingredients as input to a
function, and it will generate an output based on these ingredients. Just as
with baking, both the ingredients and the actual recipe will influence what
comes out of the recipe in the end: will it be a cake or a loaf of bread? In R,
the inputs to a function are not called ingredients, but rather *arguments*, and
the output is called the *return value* of the function. A function does not
technically have to return a value, but often does so. Functions are used to
automate more complicated sets of commands and many of them are already
predefined in R. A typical example would be the function `sqrt()`. The input
(the argument) must be a number, and the return value (in fact, the output) is
the square root of that number. Executing a function ('running it') is called
*calling* the function. An example of a function call is:

```{r}
sqrt(9)
```

Which is the same as assigning the value to a variable and then passing that
variable to the function:

```{r}
a <- 9
b <- sqrt(a)
b
```

Here, the value of `a` is given to the `sqrt()` function, the `sqrt()` function
calculates the square root, and returns the value which is then assigned to
variable `b`. This function is very simple, because it takes just one argument.

The return 'value' of a function need not be numerical (like that of `sqrt()`),
and it also does not need to be a single item: it can be a set of things, or
even a dataset, as we will see later on.

Arguments can be anything, not only numbers or filenames, but also other
objects. Exactly what each argument means differs per function, and must be
looked up in the documentation (see below). Some functions take arguments which
may either be specified by the user, or, if left out, take on a *default* value:
these are called *options*. Options are typically used to alter the way the
function operates, such as whether it ignores 'bad values', or what symbol to
use in a plot. However, if you want something specific, you can specify a value
of your choice which will be used instead of the default.

### Tab-completion

To access help about `sqrt`, we are first going to learn about tab-completion.
Type `s` and press <kbd>Tab</kbd>.

```{r, eval=FALSE}
s<tab>q
```

You can see that R gives you suggestions of what functions and variables are
available that start with the letter `s`, and thanks to RStudio they are
formatted in this nice list. There are *many* suggestions here, so let's be a
bit more specific and append a `q`, to find what we want. If we press enter or
tab again, R will insert the selected option.

You can see that R inserts a pair of parentheses together with the name of the
function. This is how the function syntax looks for R and many other
programming languages, and it means that within these parentheses, we will
specify all the arguments (the ingredients) that we want to pass to this
function.

If we press tab again, R will helpfully display all the available parameters for
this function that we can pass an argument to. The word *parameter* is used to
describe the name that the argument can be passed to. More on that later.

```{r, eval=FALSE}
sqrt(<tab>
```

There are many things in this list, but only one of them is marked in purple.
Purple here means that this list item is a parameter we can use for the function, 
while yellow means that it is a variable that we defined earlier.[^R-symbols]

[^R-symbols]: There are a few other symbols as well, all of which can be viewed
at the end of [this post about RStudio code
completion](https://support.rstudio.com/hc/en-us/articles/205273297-Code-Completion).

### Help with defined functions

To read the full help about `sqrt`, we can use the question mark, or type it
directly into the help document browser.

```{r, eval=FALSE}
?sqrt
```

As you can see, `sqrt()` takes only one argument, `x`, which needs to be a
*numerical vector*. Don't worry too much about the fact that it says *vector* here;
we will talk more about that later. Briefly, a numerical vector is one or more
numbers. In R, every number is a vector, so you don't have to do anything special to
create a vector. More on vectors later.

Let's try a function that can take multiple arguments: `round()`.

```{r, eval=FALSE}
round(<tab>)
?round
```

If we try round with a value:

```{r}
round(3.14159)
```

Here, we've called `round()` with just one argument, `3.14159`, and it has
returned the value `3`. That's because the default is to round to the nearest
whole number, or integer. If we want more digits we can pass an argument to the
`digits` parameter, to specify how many decimals we want to round to.

```{r}
round(3.14159, digits = 2)
```

So, above we pass the *argument* `2`, to the *parameter* `digits`. Knowing this
nomenclature is not essential for doing your own data analysis, but it will be
very helpful when you are reading through help documents online and in RStudio.

We can leave out the word `digits` since we know it comes as the
second parameter, after `x`.

```{r}
round(3.14159, 2)
```

As you notice, we have been leaving out `x` from the beginning. If you provide
the names for both the arguments, we can switch their order:

```{r}
round(digits = 2, x = 3.14159)
```

It's good practice to put the non-optional arguments (like the number you're
rounding) first in your function call, and to specify the names of all optional
arguments. If you don't, someone reading your code might have to look up the
definition of a function with unfamiliar arguments to understand what you're
doing.


## Writing functions

In this class, you will be working a lot with functions, especially those that
someone else has already written. When you type `sum`, `c()`, or `mean()`, you
are using a function that has been made previously and built into R. To remove
some of the magic around these functions, we will go through how to make a basic
function of our own. Let's start with a simple example where we add two numbers
together:

```{r}
add_two_numbers <- function(num1, num2) {
    return(num1 + num2)
}
add_two_numbers(4, 5)
```

As you can see, running this function on two numbers returns their sum. We
could also assign to a variable in the function and return the function.

```{r}
add_two_numbers <- function(num1, num2) {
    my_sum <- num1 + num2
    return(my_sum)
}
add_two_numbers(4, 5)
```


#### Challenge

Can you write a function that calculates the mean of 3 numbers?

```{r, include=FALSE}
mean_of_three_numbers <- function(num1, num2, num3) {
   my_sum <- num1 + num2 + num3
   my_mean <- my_sum / 3
   return(my_mean)
}
mean_of_three_numbers(2, 4, 6)
```



## Vectors and data types

A vector is the most common and basic data type in R, and is pretty much the
workhorse of R. A vector is composed by a series of values, which can be either
numbers or characters. We can assign a series of values to a vector using the
`c()` function, which stands for "concatenate (combine/connect one after
another) values into a vector" For example we can create a vector of animal
weights and assign it to a new object `weight_g`:

```{r}
weight_g <- c(50, 60, 65, 82) # Concatenate/Combine values into a vector
weight_g
```

You can also use the built-in command `seq`, to create a sequence of numbers
without typing all of them in manually.

```{r}
seq(0, 30) # This is the same as just `0:30`
seq(0, 30, 3) # Every third number
```

A vector can also contain characters:

```{r}
animals <- c('mouse', 'rat', 'dog')
animals
```

The quotes around "mouse", "rat", etc. are essential here and can be either
single or double quotes. Without the quotes R will assume there are objects
called `mouse`, `rat` and `dog`. As these objects don't exist in R's memory,
there will be an error message.

There are many functions that allow you to inspect the content of a
vector. `length()` tells you how many elements are in a particular vector:

```{r}
length(weight_g)
length(animals)
```

An important feature of a vector is that all of the elements are the same type
of data. The function `class()` indicates the class (the type of element) of an
object:

```{r}
class(weight_g)
class(animals)
```

The function `str()` provides an overview of the **str**ucture of an object
and its elements. It is a useful function when working with large and
complex objects:

```{r}
str(weight_g)
str(animals)
```

You can use the `c()` function to add other elements to your vector:

```{r}
weight_g <- c(weight_g, 90) # add to the end of the vector
weight_g <- c(30, weight_g) # add to the beginning of the vector
weight_g
```

In the first line, we take the original vector `weight_g`,
add the value `90` to the end of it, and save the result back into
`weight_g`. Then we add the value `30` to the beginning, again saving the result
back into `weight_g`.

We can do this over and over again to grow a vector, or assemble a dataset.
As we program, this may be useful to add results that we are collecting or
calculating.

An **atomic vector** is the simplest R **data type** and it is a linear vector
of a single type, e.g. all numbers. Above, we saw 2 of the 6 main **atomic
vector** types  that R uses: `"character"` and `"numeric"` (or `"double"`).
These are the basic building blocks that all R objects are built from.

Vectors are one of the many **data structures** that R uses. Other important
ones are lists (`list`), matrices (`matrix`), data frames (`data.frame`),
factors (`factor`) and arrays (`array`). In this class, we will focus on data
frames, which is most commonly used one for data analyses.

#### Challenge

We’ve seen that atomic vectors can be of type character, numeric (or double),
integer, and logical. But what happens if we try to mix these types in a single
vector? Find out by using `class` to test these examples.

```{r}
num_char <- c(1, 2, 3, 'a')
num_logical <- c(1, 2, 3, TRUE)
char_logical <- c('a', 'b', 'c', TRUE)
tricky <- c(1, 2, 3, '4')
```

```{r, include=FALSE}
# Answer
class(num_char)
class(num_logical)
class(char_logical)
class(tricky)
```

This happens because vectors can be of only one data type. Instead of
throwing an error and saying that you are trying to mix different types in
the same vector, R tries to convert (coerce) the content of this vector to
find a "common denominator". A logical can be turn into 1 or 0, and a
number can be turned into a string/character representation. It would be
difficult to do it the other way around: would 5 be TRUE or FALSE? What
number would 't' be?

In R, we call converting objects from one class into another class _coercion_.
These conversions happen according to a hierarchy, whereby some types get
preferentially coerced into other types. Can you draw a diagram that represents
the hierarchy of how these data types are coerced?

```{r, include=FALSE}
# Answer:
# logical -> numeric -> character <- logical
```

This can be important to watch for in data sets that you import.

## Subsetting vectors

If we want to extract one or several values from a vector, we must provide one
or several indices in square brackets. For instance:

```{r}
animals <- c("mouse", "rat", "dog", "cat")
animals[2]
animals[c(3, 2)]
```

We can also repeat the indices to create an object with more elements than the
original one:

```{r}
more_animals <- animals[c(1, 2, 3, 2, 1, 4)]
more_animals
```

R indices start at 1. Programming languages like Fortran, MATLAB, Julia,
and R start counting at 1, because that's what human beings typically do.
Languages in the C family (including C++, Java, Perl, and Python) count
from 0 because that was historically simpler for computers and can allow
for more elegant code.

### Conditional subsetting
Another common way of subsetting is by using a logical vector. `TRUE` will
select the element with the same index, while `FALSE` will not:

```{r}
weight_g <- c(21, 34, 39, 54, 55)
weight_g[c(TRUE, FALSE, TRUE, TRUE, FALSE)]
```

Typically, these logical vectors are not typed by hand, but are the output
of other functions or logical tests. For instance, if you wanted to select
only the values above 50:

```{r}
weight_g > 50    # will return logicals with TRUE for the indices that meet the condition
## so we can use this to select only the values above 50
weight_g[weight_g > 50]
```

We will consider conditions in more detail in the next few lectures. 

### Strings (character vectors)

Just a small note about character vectors, also called `strings`. There are
built-in packages for subsetting them that we'll learn about later. They can be
particularly relevant for ecological and genomic data because important data can be
nested in complicated strings of text (ex: extracting only the observations
that occurred in wet habitats from a column of habitat descriptions or only
genes with functions related to drought tolerance).

```{r}
string1 <- "This is a string" # you can include spaces between your quotes
string2 <- c(string1, "so is this") # concatenate with another string
string2[2] # can access the second string via subsetting

# Playing a bit with declaring variables
"You can include 'quotes' in a string"
string3 <- 'You can include "quotes" in a string'
string3
"You can include \"matching quotes\" if you 'escape' them with a backslash (\\)"
```

## Missing data

As R was designed to analyze datasets, it includes the concept of missing
data (which is uncommon in other programming languages). Missing data are
represented in vectors as `NA`.

When doing operations on numbers, most functions will return `NA` if the
data you are working with include missing values. This feature makes it
harder to overlook the cases where you are dealing with missing data. You
can add the argument `na.rm = TRUE` to calculate the result while ignoring
the missing values.

```{r}
heights <- c(2, 4, 4, NA, 6)
mean(heights)
max(heights)
mean(heights, na.rm = TRUE)
max(heights, na.rm = TRUE)
```

```{r}
## Extract those elements which are not missing values.
heights[!is.na(heights)]

## Returns the object with incomplete cases removed. The returned object is an atomic vector of type `"numeric"` (or `"double"`).
na.omit(heights)

## Extract those elements which are complete cases. The returned object is an atomic vector of type `"numeric"` (or `"double"`).
heights[complete.cases(heights)]
```

Recall that you can use the `class()` function to find the type of your atomic vector.

#### Challenge

1. Using this vector of length measurements, create a new vector with the NAs
removed.

```{r}
lengths <- c(10, 24, NA, 18, NA, 20)
```

```{r, include=FALSE}
# Answer
na.omit(lengths)
```

2. Use the function `median()` to calculate the median of the `lengths` vector.

```{r, include=FALSE}
# Answer
median(lengths, na.rm = TRUE)
```

## Loops and vectorization

Loops, specifically for-loops, are essential to programming in general. However, in R, 
you should avoid them as often as possible because there are more efficient ways of
doing things that you should use instead. It is still important that you
understand the concept of loops and you might also use them in some of your own
functions if there is no vectorized way of going about what you want to do.

You can think of a for-loop as: "for each number contained in a list/vector,
perform this operation" and the syntax basically says the same thing:

```{r}
v <- c(2, 4, 6)
for (num in v) {
    print(num)
}
```

Instead of printing out every number to the console, we could also add numbers
cumulatively, to calculate the sum of all the numbers in the vector:

```{r}
# To increment `w` each time, we must first create the variable,
# which we do by setting `w <- 0`, referred to as initializing.
# This also ensures that `w` is zero at the start of the loop and
# doesn't retain the value from last time we ran this code.
w <- 0
for (num in v) {
    w <- w + num
}
w
```

If we put what we just did inside a function, we have essentially recreated the
`sum` function in R.

```{r}
my_sum <- function(input_vector) {
    vector_sum <- 0
    for (num in input_vector){
        vector_sum <- vector_sum + num
    }
    return(vector_sum)
}

my_sum(v)
```

Although this gives us the same output as the built-in function `sum`, the
built-in function has many more optimizations so it is much faster than our
function. In R, it is always faster to try to find a way of doing things without
writing a loop yourself. When you are reading about R, you might see suggestions
that you should try to *vectorize* your code to make it faster. What people are
referring to, is that you should not write for loops in R and instead use the
ready-made functions that are much more efficient in working with vectors and
essentially performs operations on entire vector at once instead of one number
at a time. Conceptually, loops operate on one element at a time while vectorized
code operates on all elements of a vector at once.

## Dataset background

Today, we will be working with real data from a longitudinal study of the
species abundance in the Chihuahuan desert ecosystem near Portal, Arizona, USA.
This study includes observations of plants, ants, and rodents from 1977 - 2002,
and has been used in over 100 publications. More information is available in
[the abstract of this paper from 2009](
https://onlinelibrary.wiley.com/doi/10.1890/08-1222.1/full). There are several
datasets available related to this study, and we will be working with datasets
that have been preprocessed by the [Data
Carpentry](https://www.datacarpentry.org) to facilitate teaching. These are made
available online as _The Portal Project Teaching Database_, both at the [Data
Carpentry website](https://datacarpentry.org/ecology-workshop/data.html), and on
[Figshare](https://figshare.com/articles/Portal_Project_Teaching_Database/1314459/6).
Figshare is a great place to publish data, code, figures, and more openly to
make them available for other researchers and to communicate findings that are
not part of a longer paper.

### Presentation of the survey data

We are studying the species and weight of animals caught in plots in our study
area. The dataset is stored as a comma separated value (CSV) file. Each row
holds information for a single animal, and the columns represent:

| Column           | Description                        |
|------------------|------------------------------------|
| record_id        | unique id for the observation      |
| month            | month of observation               |
| day              | day of observation                 |
| year             | year of observation                |
| plot_id          | ID of a particular plot            |
| species_id       | 2-letter code                      |
| sex              | sex of animal ("M", "F")           |
| hindfoot_length  | length of the hindfoot in mm       |
| weight           | weight of the animal in grams      |
| genus            | genus of animal                    |
| species          | species of animal                  |
| taxa             | e.g. rodent, reptile, bird, rabbit |
| plot_type        | type of plot                       |

To read the data into R, we are going to use a function called `read_csv`. This
function is contained in an R-package called
[`readr`](https://readr.tidyverse.org/). R-packages are a bit like browser
extensions; they are not essential, but can provide nifty functionality. We will
go through R-packages in general and which ones are good for data analyses. One useful option that `read_csv`
includes, is the ability to read a CSV file directly from a URL, without
downloading it in a separate step:

```{r, eval=FALSE}
surveys <- readr::read_csv('https://ndownloader.figshare.com/files/2292169')
```

However, it is often a good idea to download the data first, so you have a copy
stored locally on your computer in case you want to do some offline analyses, or
the online version of the file changes or the file is taken down. You can either
download the data manually or from within R:

```{r, eval=FALSE}
download.file("https://ndownloader.figshare.com/files/2292169",
              "data/portal_data.csv") # Saves to current directory with this name
```

The data is read in by specifying its local path.

```{r}
surveys <- readr::read_csv('data/portal_data.csv')
```

This statement produces some output regarding which data type it found in each 
column. If we want to check this in more detail, we can print the variable's 
value: `surveys`.

```{r}
surveys
```

This displays a nice tabular view of the data, which also includes
pagination when there are many rows and we can click the arrow to view all
the columns. Technically, this object is actually a `tibble` rather than a
data frame, as indicated in the output. The reason for this is that
`read_csv` automatically converts the data into to a `tibble` when loading
it. Since a `tibble` is just a data frame with some convenient extra
functionality, we will use these words interchangeably from now on.

If we just want to glance at how the data frame looks, it is sufficient to
display only the top (the first 6 lines) using the function `head()`:

```{r}
head(surveys)
```

## What are data frames?

Data frames are the _de facto_ data structure for most tabular data, and what we
use for statistics and plotting. A data frame can be created by hand, but most
commonly they are generated by the function `read_csv()`; in other words, when
importing spreadsheets from your hard drive (or the web).

A data frame is a representation of data in the format of a table where the
columns are vectors that all have the same length. Because the columns are
vectors, they all contain the same type of data as we discussed in last class
(e.g., characters, integers, factors). We can see this when inspecting the
structure of a data frame with the function `str()`:

```{r}
str(surveys)
```

Integer refers to a whole number, such as 1, 2, 3, 4, etc. Numbers with
decimals, 1.0, 2.4, 3.333, are referred to as floats. Factors are used to
represent categorical data. Factors can be ordered or unordered, and
understanding them is necessary for statistical analysis and for plotting.
Factors are stored as integers, and have labels (text) associated with these
unique integers. While factors look (and often behave) like character vectors,
they are actually integers under the hood, and you need to be careful when
treating them like strings.

### Inspecting `data.frame` objects

We already saw how the functions `head()` and `str()` can be useful to check the
content and the structure of a data frame. Here is a non-exhaustive list of
functions to get a sense of the content/structure of the data. Let's try them
out!

* Size:
    * `dim(surveys)` - returns a vector with the number of rows in the first element
      and the number of columns as the second element (the dimensions of the object)
    * `nrow(surveys)` - returns the number of rows
    * `ncol(surveys)` - returns the number of columns

* Content:
    * `head(surveys)` - shows the first 6 rows
    * `tail(surveys)` - shows the last 6 rows

* Names:
    * `names(surveys)` - returns the column names (synonym of `colnames()` for `data.frame`
      objects)
    * `rownames(surveys)` - returns the row names

* Summary:
    * `str(surveys)` - structure of the object and information about the class,
      length, and content of  each column
    * `summary(surveys)` - summary statistics for each column

Note: most of these functions are "generic", they can be used on other types of
objects besides `data.frame`.

#### Challenge

Based on the output of `str(surveys)`, can you answer the following questions?

* What is the class of the object `surveys`?
* How many rows and how many columns are in this object?
* How many species have been recorded during these surveys?

```{r, include=FALSE}
## Answers
##
## * class: data frame
## * how many rows: 34786,  how many columns: 13
## * how many species: 48
```

### Indexing and subsetting data frames

Our survey data frame has rows and columns (it has 2 dimensions). If we want to
extract some specific data from it, we need to specify the "coordinates" we
want from it. Row numbers come first, followed by column numbers. When
indexing, base R data frames return a different format depending on how we
index the data (i.e. either a vector or a data frame), but with enhanced data
frames, `tibbles`, the returned object is almost always a data frame.

```{r}
surveys[1, 1]   # first element in the first column of the data frame
surveys[1, 6]   # first element in the 6th column
surveys[, 1]    # first column in the data frame
surveys[1]      # first column in the data frame
surveys[1:3, 7] # first three elements in the 7th column
surveys[3, ]    # the 3rd element for all columns
surveys[1:6, ]  # equivalent to head(surveys)
```

`:` is a special operator that creates numeric vectors of integers in
increasing or decreasing order; test `1:10` and `10:1` for instance. This works
similarly to `seq`, which we looked at earlier in class:

```{r}
0:10
seq(0, 10)

# We can test if all elements are the same
0:10 == seq(0,10)
all(0:10 == seq(0,10))
```

You can also exclude certain parts of a data frame using the "`-`" sign:

```{r}
surveys[,-1]    # All columns, except the first
surveys[-c(7:34786),] # Equivalent to head(surveys)
```

As well as using numeric values to subset a `data.frame` (or `matrix`), columns
can be called by name, using one of the four following notations: <!-- Not sure how important it is to learn the difference vs just teaching the preferred way with the footnote that there are other ways also. -->

```{r}
surveys["species_id"]       # Result is a data.frame
surveys[, "species_id"]     # Result is a data.frame
```

For our purposes, these notations are equivalent. RStudio knows about
the columns in your data frame, so you can take advantage of the autocompletion
feature to get the full and correct column name.

Another syntax that is often used to specify column names is `$`. In this case,
the returned object is actually a vector. We will not go into detail about this,
but since it is such common usage, it is good to be aware of this.

```{r}
# We use `head()` since the output from vectors are not automatically cut off
# and we don't want to clutter the screen with all the `species_id` values
head(surveys$species_id)          # Result is a vector
```

#### Challenge

1. Create a `data.frame` (`surveys_200`) containing only the observations from
   row 200 of the `surveys` dataset.

2. Notice how `nrow()` gave you the number of rows in a `data.frame`?

     * Use that number to pull out just that last row in the data frame.
     * Compare that with what you see as the last row using `tail()` to make
       sure it's meeting expectations.
     * Pull out that last row using `nrow()` instead of the row number.
     * Create a new data frame object (`surveys_last`) from that last row.

3. Use `nrow()` to extract the row that is in the middle of the data
   frame. Store the content of this row in an object named `surveys_middle`.

4. Combine `nrow()` with the `-` notation above to reproduce the behavior of
   `head(surveys)` keeping just the first through 6th rows of the surveys
   dataset.

```{r, include=FALSE}
## Answers
surveys_200 <- surveys[200, ]
surveys_last <- surveys[nrow(surveys), ]
surveys_middle <- surveys[nrow(surveys)/2, ]
surveys_head <- surveys[-c(7:nrow(surveys)),]
```

## Exporting data

As you begin to play with your raw data, you may want to export these new,
processed, datasets to share them with your collaborators or for archival.
Similar to the `read_csv()` function used for reading CSV files into R,
there is a `write_csv()` function that generates CSV files from data
frames.

Manually create a new folder called "data-processed" in your directory. Alternatively, get R to help you with it.

```{r, eval=FALSE}
dir.create("Processed data")
```

We are going to prepare a cleaned up version of the data without NAs. 

```{r}
# Note that this omits observations with NA in *any* column.
# There is no way to control which columns to use.
surveys_complete_naomit <- na.omit(surveys)

# Compare the dimensions of the original and the cleaned data frame
dim(surveys)
dim(surveys_complete_naomit)
```

Now that our dataset is ready, we can save it as a CSV file in our `Processed data`
folder.

```{r, eval=FALSE}
# To save to current directory
write_csv(surveys_complete_naomit, "surveys_complete_naomit.csv")

# To save to newly created directory
write_csv(surveys_complete_naomit, 
          file.path("~/Processed data", "surveys_complete_naomit.csv"))
```

Next lecture, we're going to discuss collaboration with GitHub and go over
an intro to the command line.

