# Manipulating and analyzing data

## Lesson preamble

> ### Learning objectives
>
> -   Describe what a data frame is.
> -   Load external data from a .csv file into a data frame in R.
> -   Understand the purpose of the **`dplyr`** package.
> -   Learn to use data wrangling commands `select`, `filter`, `%>%,`
>     and `mutate` from the **`dplyr`** package.
> -   Understand the split-apply-combine concept for data analysis.
> -   Use `summarize`, `group_by`, and `tally` to split a data frame
>     into groups of observations, apply a summary statistics for each
>     group, and then combine the results.
> -   Learn to switch between long and wide format
>
> ### Lesson outline
>
> -   Getting set up with R Markdown and R packages for data analytics
> -   Data set background (10 min)
> -   What are data frames? (10 min)
> -   Split-apply-combine techniques in **`dplyr`** (25 min)
> -   Using `group_by` and `tally` to summarize categorical data (20
>     mins)
> -   Reshaping data (15 mins)

------------------------------------------------------------------------

## Set up

To do this lecture, we'll need to be able to create and knit R Markdown
documents, and analyze data using functions contained in the `tidyverse`
package. If you haven't done either of these tasks before, navigate to
the course notes for [Lecture 1](lec01-downloading-r.qmd) and follow the
instructions there first.

When we make Markdown documents, we want them to be shareable and
reproducible, so it's good practice to include code to install and load
any packages the code requires, even if we've already done it ourselves.
While packages only need to be installed once (they'll still be there
even if RStudio is closed and reopened again or if R is accessed outside
of RStudio), someone we share our code with might have never used them
before. Or, if you move from using your code on one computer to another,
or one computer to the POSIT Cloud, you'll need to re-install.

We can add code install new packages using the function
`install.packages()`. We'll pass `eval=FALSE` to *knitr* at the top of
our code chunk to make sure that the chunk won't be *eval*uated when we
knit the document, since we don't really need to install it each time,
and this option can be changed by another user. You can find other
[possible options](https://yihui.org/knitr/options/) to pass that can be
helpful for formatting your output document.

```{r, eval=FALSE}
install.packages('tidyverse')
```

The two `tidyverse` packages we will be using the most frequently in
this course is `dplyr` and `ggplot2`. `dplyr` is great for data
wrangling (Lecture 3) and `ggplot2` makes killer plots (Lecture 4).

To use functions in the `dplyr` package, type `dplyr::` and then the
function name.

```{r}
dplyr::glimpse(cars) # `glimpse` is similar to `str`
# cars is a built-in data set
```

Since we will be using this package a lot, it would be a little annoying
to have to type `dplyr::` every time. We can bypass this step by loading
the package into our current environment. Think of this is "opening" the
package for your work session.

```{r}
# We could also do `library(dplyr)`, but we need the rest of the
# tidyverse packages later, so we might as well import the entire collection.
library(tidyverse)

glimpse(cars)
```

This needs to be done once for every new R session, and so it is common
practice to keep a list of all the packages used at the top of your
script or notebook for convenience and load all of it at start up.

That's a lot of red though! What are these warning signs and checks?

All the warning signs indicate are the version of R that they were built
under. They can frequently be ignored unless your version of R is so old
that the packages can no longer be run on R! Note that packages are
frequently updated, and functions may become deprecated.

Next, the warning shows you all the packages that were successfully
installed.

Finally, there are some conflicts! All this means is that there are
multiple functions with the same name that may do different things. R
prioritizes functions from certain packages over others. So, in this
case, the `filter()` function from `dplyr` will take precedent over the
`filter()` function from the `stats` package. If you want to use the
latter, use double colons `::` to indicate that you are calling a
function from a certain package:

## Dataset background

Today, we will be working with real data from a longitudinal study of
the species abundance in the Chihuahuan desert ecosystem near Portal,
Arizona, USA. This study includes observations of plants, ants, and
rodents from 1977 - 2002, and has been used in over 100 publications.
More information is available in [the abstract of this paper from
2009](https://onlinelibrary.wiley.com/doi/10.1890/08-1222.1/full). There
are several datasets available related to this study, and we will be
working with datasets that have been preprocessed by the [Data
Carpentry](https://www.datacarpentry.org) to facilitate teaching. These
are made available online as *The Portal Project Teaching Database*,
both at the [Data Carpentry
website](https://datacarpentry.org/ecology-workshop/data.html), and on
[Figshare](https://figshare.com/articles/Portal_Project_Teaching_Database/1314459/6).
Figshare is a great place to publish data, code, figures, and more
openly to make them available for other researchers and to communicate
findings that are not part of a longer paper.

### Presentation of the survey data

We are studying the species and weight of animals caught in plots in our
study area. The dataset is stored as a comma separated value (CSV) file.
Each row holds information for a single animal, and the columns
represent:

| Column          | Description                        |
|-----------------|------------------------------------|
| record_id       | unique id for the observation      |
| month           | month of observation               |
| day             | day of observation                 |
| year            | year of observation                |
| plot_id         | ID of a particular plot            |
| species_id      | 2-letter code                      |
| sex             | sex of animal ("M", "F")           |
| hindfoot_length | length of the hindfoot in mm       |
| weight          | weight of the animal in grams      |
| genus           | genus of animal                    |
| species         | species of animal                  |
| taxa            | e.g. rodent, reptile, bird, rabbit |
| plot_type       | type of plot                       |

To read the data into R, we are going to use a function called
`read_csv`. This function is contained in an R-package called
[`readr`](https://readr.tidyverse.org/). R-packages are a bit like
browser extensions; they are not essential, but can provide nifty
functionality. We will go through R-packages in general and which ones
are good for data analyses. One useful option that `read_csv` includes,
is the ability to read a CSV file directly from a URL, without
downloading it in a separate step:

```{r}
library(readr)
```

```{r, eval=FALSE}
surveys <- readr::read_csv('https://ndownloader.figshare.com/files/2292169')
```

However, it is often a good idea to download the data first, so you have
a copy stored locally on your computer in case you want to do some
offline analyses, or the online version of the file changes or the file
is taken down. You can either download the data manually or from within
R:

```{r}
download.file("https://ndownloader.figshare.com/files/2292169",
              "data/portal_data.csv") # Saves to current directory with this name
```

The data is read in by specifying its local path.

```{r}
surveys <- readr::read_csv('data/portal_data.csv')
```

This statement produces some output regarding which data type it found
in each column. If we want to check this in more detail, we can print
the variable's value: `surveys`.

```{r}
surveys
```

This displays a nice tabular view of the data, which also includes
pagination when there are many rows and we can click the arrow to view
all the columns. Technically, this object is actually a `tibble` rather
than a data frame, as indicated in the output. The reason for this is
that `read_csv` automatically converts the data into to a `tibble` when
loading it. Since a `tibble` is just a data frame with some convenient
extra functionality, we will use these words interchangeably from now
on.

If we just want to glance at how the data frame looks, it is sufficient
to display only the top (the first 6 lines) using the function `head()`:

```{r}
head(surveys)
```

## What are data frames?

Data frames are the *de facto* data structure for most tabular data, and
what we use for statistics and plotting. A data frame can be created by
hand, but most commonly they are generated by the function `read_csv()`;
in other words, when importing spreadsheets from your hard drive (or the
web).

A data frame is a representation of data in the format of a table where
the columns are vectors that all have the same length. Because the
columns are vectors, they all contain the same type of data as we
discussed in last class (e.g., characters, integers, factors). We can
see this when inspecting the structure of a data frame with the function
`str()`:

```{r}
str(surveys)
```

Integer refers to a whole number, such as 1, 2, 3, 4, etc. Numbers with
decimals, 1.0, 2.4, 3.333, are referred to as floats. Factors are used
to represent categorical data. Factors can be ordered or unordered, and
understanding them is necessary for statistical analysis and for
plotting. Factors are stored as integers, and have labels (text)
associated with these unique integers. While factors look (and often
behave) like character vectors, they are actually integers under the
hood, and you need to be careful when treating them like strings.

### Inspecting `data.frame` objects

We already saw how the functions `head()` and `str()` can be useful to
check the content and the structure of a data frame. Here is a
non-exhaustive list of functions to get a sense of the content/structure
of the data. Let's try them out!

-   Size:
    -   `dim(surveys)` - returns a vector with the number of rows in the
        first element and the number of columns as the second element
        (the dimensions of the object)
    -   `nrow(surveys)` - returns the number of rows
    -   `ncol(surveys)` - returns the number of columns
-   Content:
    -   `head(surveys)` - shows the first 6 rows
    -   `tail(surveys)` - shows the last 6 rows
-   Names:
    -   `names(surveys)` - returns the column names (synonym of
        `colnames()` for `data.frame` objects)
    -   `rownames(surveys)` - returns the row names
-   Summary:
    -   `str(surveys)` - structure of the object and information about
        the class, length, and content of each column
    -   `summary(surveys)` - summary statistics for each column

Note: most of these functions are "generic", they can be used on other
types of objects besides `data.frame`.

#### Challenge

Based on the output of `str(surveys)`, can you answer the following
questions?

-   What is the class of the object `surveys`?
-   How many rows and how many columns are in this object?
-   How many species have been recorded during these surveys?

```{r, include=FALSE}
## Answers
##
## * class: data frame
## * how many rows: 34786,  how many columns: 13
## * how many species: 48
```

### Indexing and subsetting data frames

Our survey data frame has rows and columns (it has 2 dimensions). If we
want to extract some specific data from it, we need to specify the
"coordinates" we want from it. Row numbers come first, followed by
column numbers. When indexing, base R data frames return a different
format depending on how we index the data (i.e. either a vector or a
data frame), but with enhanced data frames, `tibbles`, the returned
object is almost always a data frame.

```{r}
surveys[1, 1]   # first element in the first column of the data frame
surveys[1, 6]   # first element in the 6th column
surveys[, 1]    # first column in the data frame
surveys[1]      # first column in the data frame
surveys[1:3, 7] # first three elements in the 7th column
surveys[3, ]    # the 3rd element for all columns
surveys[1:6, ]  # equivalent to head(surveys)
```

`:` is a special operator that creates numeric vectors of integers in
increasing or decreasing order; test `1:10` and `10:1` for instance.
This works similarly to `seq`, which we looked at earlier in class:

```{r}
0:10
seq(0, 10)

# We can test if all elements are the same
0:10 == seq(0,10)
all(0:10 == seq(0,10))
```

You can also exclude certain parts of a data frame using the "`-`" sign:

```{r}
surveys[,-1]    # All columns, except the first
surveys[-c(7:34786),] # Equivalent to head(surveys)
```

As well as using numeric values to subset a `data.frame` (or `matrix`),
columns can be called by name, using one of the four following
notations:
<!-- Not sure how important it is to learn the difference vs just teaching the preferred way with the footnote that there are other ways also. -->

```{r}
surveys["species_id"]       # Result is a data.frame
surveys[, "species_id"]     # Result is a data.frame
```

For our purposes, these notations are equivalent. RStudio knows about
the columns in your data frame, so you can take advantage of the
autocompletion feature to get the full and correct column name.

Another syntax that is often used to specify column names is `$`. In
this case, the returned object is actually a vector. We will not go into
detail about this, but since it is such common usage, it is good to be
aware of this.

```{r}
# We use `head()` since the output from vectors are not automatically cut off
# and we don't want to clutter the screen with all the `species_id` values
head(surveys$species_id)          # Result is a vector
```

#### Challenge

1.  Create a `data.frame` (`surveys_200`) containing only the
    observations from row 200 of the `surveys` dataset.

2.  Notice how `nrow()` gave you the number of rows in a `data.frame`?

    -   Use that number to pull out just that last row in the data
        frame.
    -   Compare that with what you see as the last row using `tail()` to
        make sure it's meeting expectations.
    -   Pull out that last row using `nrow()` instead of the row number.
    -   Create a new data frame object (`surveys_last`) from that last
        row.

3.  Use `nrow()` to extract the row that is in the middle of the data
    frame. Store the content of this row in an object named
    `surveys_middle`.

4.  Combine `nrow()` with the `-` notation above to reproduce the
    behavior of `head(surveys)` keeping just the first through 6th rows
    of the surveys dataset.

```{r, include=FALSE}
## Answers
surveys_200 <- surveys[200, ]
surveys_last <- surveys[nrow(surveys), ]
surveys_middle <- surveys[nrow(surveys)/2, ]
surveys_head <- surveys[-c(7:nrow(surveys)),]
```

## Exporting data

As you begin to play with your raw data, you may want to export these
new, processed, datasets to share them with your collaborators or for
archival. Similar to the `read_csv()` function used for reading CSV
files into R, there is a `write_csv()` function that generates CSV files
from data frames.

Manually create a new folder called "data-processed" in your directory.
Alternatively, get R to help you with it.

```{r, eval=FALSE}
dir.create("Processed data")
```

We are going to prepare a cleaned up version of the data without NAs.

```{r}
# Note that this omits observations with NA in *any* column.
# There is no way to control which columns to use.
surveys_complete_naomit <- na.omit(surveys)

# Compare the dimensions of the original and the cleaned data frame
dim(surveys)
dim(surveys_complete_naomit)
```

Now that our dataset is ready, we can save it as a CSV file in our
`Processed data` folder.

```{r, eval=FALSE}
# To save to current directory
write_csv(surveys_complete_naomit, "surveys_complete_naomit.csv")

# To save to newly created directory
write_csv(surveys_complete_naomit, 
          file.path("Processed data", 
                    "surveys_complete_naomit.csv"))
```

## 

## Data wrangling with dplyr

Wrangling here is used in the sense of maneuvering, managing,
controlling, and turning your data upside down and inside out to look at
it from different angles in order to understand it. The package
**`dplyr`** provides easy tools for the most common data manipulation
tasks. It is built to work directly with data frames, with many common
tasks optimized by being written in a compiled language (C++), this
means that many operations run much faster than similar tools in R. An
additional feature is the ability to work directly with data stored in
an external database, such as SQL-databases. The ability to work with
databases is great because you are able to work with much bigger
datasets (100s of GB) than your computer could normally handle. We will
not talk in detail about this in class, but there are great resources
online to learn more (e.g. [this lecture from Data
Carpentry](https://datacarpentry.org/R-ecology-lesson/05-r-and-databases.html)).

### Selecting columns and filtering rows

We're going to learn some of the most common **`dplyr`** functions:
`select()`, `filter()`, `mutate()`, `group_by()`, and `summarise()`. To
select columns of a data frame, use `select()`. The first argument to
this function is the data frame (`surveys`), and the subsequent
arguments are the columns to keep. Note that we don't need quotation
marks around the column names here like with did with base R. You do
still need quotation marks around strings, though!

```{r, echo=FALSE}
select(surveys, plot_id, species_id, weight, year)
```

To choose rows based on a specific criteria, use `filter()`:

```{r, echo=FALSE}
filter(surveys, year == 1995)
```

#### An aside on conditionals

Note that to check for equality, R requires *two* equal signs (`==`).
This is to prevent confusion with object assignment, since otherwise
`year = 1995` might be interpreted as 'set the `year` parameter to
`1995`', which is not what `filter` does!

Basic conditionals in R are broadly similar to how they're already
expressed mathematically:

```{r}
2 < 3
5 > 9
```

However, there are a few idiosyncrasies to be mindful of for other
conditionals:

```{r}
2 != 3 # not equal
2 <= 3 # less than or equal to
5 >= 9 # greater than or equal to
```

Finally, the `%in%` operator is used to check for membership:

```{r}
2 %in% c(2, 3, 4) # check whether 2 in c(2, 3, 4)
```

All of the above conditionals are compatible with `filter`, with the key
difference being that `filter` expects column names as part of
conditional statements instead of individual numbers.

### Chaining functions together using pipes

But what if you wanted to select and filter at the same time? There are
three ways to do this: use intermediate steps, nested functions, or
pipes. With *intermediate steps*, you essentially create a temporary
data frame and use that as input to the next function. This can clutter
up your workspace with lots of objects:

```{r, echo = FALSE}
temp_df <- select(surveys, plot_id, species_id, weight, year)
filter(temp_df, year == 1995)
```

You can also *nest functions* (i.e. one function inside of another).
This is handy, but can be difficult to read if too many functions are
nested as things are evaluated from the inside out. Readability can be
mildly improved by enabling "rainbow parentheses" (open settings \> Code
\> Display and check rainbow parentheses), but it's still basically
impossible to document and effectively convey your work with this
method.

```{r}
filter(select(surveys, plot_id, species_id, weight, year), year == 1995)
```

The last option, *pipes*, are a fairly recent addition to R. Pipes let
you take the output of one function and send it directly to the next,
which is useful when you need to do many things to the same dataset.
Pipes in R look like `%>%` and are made available via the `magrittr`
package that also is included in the `tidyverse`. If you use RStudio,
you can type the pipe with <kbd>Ctrl/Cmd</kbd> + <kbd>Shift</kbd> +
<kbd>M</kbd>.

```{r}
surveys %>% 
    select(., plot_id, species_id, weight, year) %>% 
    filter(., year == 1995)
```

The `.` refers to the object that is passed from the previous line. In
this example, the data frame `surveys` is passed to the `.` in the
`select()` statement. Then, the modified data frame which is the result
of the `select()` operation, is passed to the `.` in the filter()
statement. Put more simply: whatever was the result from the line above
the current line, will be used in the current line.

Since it gets a bit tedious to write out all the dots, **`dplyr`**
allows for them to be omitted. By default, the pipe will pass its input
to the *first* argument of the right hand side function; in `dplyr`, the
first argument is always a data frame. The chunk below gives the same
output as the one above:

```{r}
surveys %>% 
    select(plot_id, species_id, weight, year) %>% 
    filter(year == 1995)
```

Another example:

```{r}
surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)
```

In the above code, we use the pipe to send the `surveys` dataset first
through `filter()` to keep rows where `weight` is less than 5, then
through `select()` to keep only the `species_id`, `sex`, and `weight`
columns. Since `%>%` takes the object on its left and passes it as the
first argument to the function on its right, we don't need to explicitly
include it as an argument to the `filter()` and `select()` functions
anymore.

If this runs off your screen and you just want to see the first few
rows, you can use a pipe to view the `head()` of the data. (Pipes work
with non-**`dplyr`** functions, too, as long as either the **`dplyr`**
or `magrittr` package is loaded).

```{r}
surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight) %>% 
  head()
```

If we wanted to create a new object with this smaller version of the
data, we could do so by assigning it a new name:

```{r}
surveys_sml <- surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)

surveys_sml
```

Note that the final data frame is the leftmost part of this expression.

A single expression can also be used to filter for several criteria,
either matching *all* criteria (`&`) or *any* criteria (`|`):

```{r}
surveys %>% 
    filter(taxa == 'Rodent' & sex == 'F') %>% 
    select(sex, taxa)
```

```{r}
surveys %>% 
    filter(species == 'clarki' | species == 'leucophrys') %>% 
    select(species, taxa)
```

#### Challenge

Using pipes, subset the `survey` data to include individuals collected
before 1995 and retain only the columns `year`, `sex`, and `weight`.

```{r, include=FALSE}
## Answer
surveys %>%
    filter(year < 1995) %>%
    select(year, sex, weight)
```

### Creating new columns with mutate

Frequently, you'll want to create new columns based on the values in
existing columns. For instance, you might want to do unit conversions,
or find the ratio of values in two columns. For this we'll use
`mutate()`.

To create a new column of weight in kg:

```{r}
surveys %>%
    mutate(weight_kg = weight / 1000)
```

You can also create a second new column based on the first new column
within the same call of `mutate()`:

```{r}
surveys %>%
    mutate(weight_kg = weight / 1000,
           weight_kg2 = weight_kg * 2)
```

The first few rows of the output are full of `NA`s, so if we wanted to
remove those we could insert a `filter()` in the chain:

```{r}
surveys %>%
    filter(!is.na(weight)) %>%
    mutate(weight_kg = weight / 1000)
```

`is.na()` is a function that determines whether something is an `NA`.
The `!` symbol negates the result, so we're asking for everything that
*is not* an `NA`.

#### Challenge

Create a new data frame from the `surveys` data that meets the following
criteria: contains only the `species_id` column and a new column called
`hindfoot_half` containing values that are half the `hindfoot_length`
values. In this `hindfoot_half` column, there are no `NA`s and all
values are less than 30.

**Hint**: think about how the commands should be ordered to produce this
data frame!

```{r, include=FALSE}
## Answer
surveys_hindfoot_half <- surveys %>%
    filter(!is.na(hindfoot_length)) %>%
    mutate(hindfoot_half = hindfoot_length / 2) %>%
    filter(hindfoot_half < 30) %>%
    select(species_id, hindfoot_half)
```

## Split-apply-combine techniques in dplyr

Many data analysis tasks can be approached using the
*split-apply-combine* paradigm: split the data into groups, apply some
analysis to each group, and then combine the results.

**`dplyr`** facilitates this workflow through the use of `group_by()` to
split data and `summarize()`, which collapses each group into a
single-row summary of that group. The arguments to `group_by()` are the
column names that contain the **categorical** variables for which you
want to calculate the summary statistics. Let's view the mean `weight`
by sex.

```{r}
surveys %>%
    group_by(sex) %>%
    summarize(mean_weight = mean(weight))
```

The mean weights become `NA` since there are individual observations
that are `NA`. Let's remove those observations.

```{r}
surveys %>%
    filter(!is.na(weight)) %>%
    group_by(sex) %>%
    summarize(mean_weight = mean(weight))
```

There is one row here that is neither male nor female, these are
observations where the animal escaped before the sex could not be
determined. Let's remove those as well.

```{r}
surveys %>%
    filter(!is.na(weight) & !is.na(sex)) %>%
    group_by(sex) %>%
    summarize(mean_weight = mean(weight))
```

You can also group by multiple columns:

```{r}
surveys %>%
    filter(!is.na(weight) & !is.na(sex)) %>%
    group_by(genus, sex) %>%
    summarize(mean_weight = mean(weight))
```

Since we will use the same filtered and grouped data frame in multiple
code chunks below, we could assign this subset of the data to a new
variable and use this variable in the subsequent code chunks instead of
typing out the functions each time.

```{r}
filtered_surveys <- surveys %>%
    filter(!is.na(weight) & !is.na(sex)) %>%
    group_by(genus, sex)
```

If you want to display more data, you can use the `print()` function at
the end of your chain with the argument `n` specifying the number of
rows to display.

```{r}
filtered_surveys %>%
    summarize(mean_weight = mean(weight)) %>%
    print(n = 15) # Will change the knitted output, not the notebook
```

Once the data are grouped, you can also summarize multiple variables at
the same time. For instance, we could add a column indicating the
minimum weight for each species for each sex:

```{r}
filtered_surveys %>%
    summarize(mean_weight = mean(weight),
              min_weight = min(weight))
```

#### Challenge

1.  Use `group_by()` and `summarize()` to find the mean, min, and max
    hindfoot length for each species.

2.  What was the heaviest animal measured in each year? Return the
    columns `year`, `genus`, `species`, and `weight`.

```{r, include=FALSE}
## Answer 1
surveys %>%
    filter(!is.na(hindfoot_length)) %>%
    group_by(species) %>%
    summarize(
        mean_hindfoot_length = mean(hindfoot_length),
        min_hindfoot_length = min(hindfoot_length),
        max_hindfoot_length = max(hindfoot_length)
    ) %>%
    print(n = 15)
```

```{r, include=FALSE}
## Answer 2
surveys %>%
    filter(!is.na(weight)) %>%
    group_by(year) %>%
    filter(weight == max(weight)) %>% # This is going to compare to the max weight within each group
    select(year, genus, species, weight) %>%
    arrange(year) %>%
    print(n = 15)
```

### Using tally to summarize categorical data

When working with data, it is also common to want to know the number of
observations found for each factor or combination of factors. For this,
**`dplyr`** provides `tally()`. For example, if we want to group by taxa
and find the number of observations for each taxa, we would do:

```{r}
surveys %>%
    group_by(taxa) %>%
    tally()
```

We can also use `tally()` when grouping on multiple variables:

```{r}
surveys %>%
    group_by(taxa, sex) %>%
    tally()
```

Here, `tally()` is the action applied to the groups created by
`group_by()` and counts the total number of records for each category.

If there are many groups, `tally()` is not that useful on its own. For
example, when we want to view the five most abundant species among the
observations:

```{r}
surveys %>%
    group_by(species) %>%
    tally()
```

Since there are 40 rows in this output, we would like to order the table
to display the most abundant species first. In `dplyr`, we say that we
want to `arrange()` the data.

```{r}
surveys %>%
    group_by(species) %>%
    tally() %>%
    arrange(n)
```

Still not that useful. Since we are interested in the most abundant
species, we want to display those with the highest count first, in other
words, we want to arrange the column `n` in descending order:

```{r}
surveys %>%
    group_by(species) %>%
    tally() %>%
    arrange(desc(n)) %>%
    head(5)
```

If we want to include more attributes about these species, we can
include these in the call to `group_by()`:

```{r}
surveys %>%
    group_by(species, taxa, genus) %>%
    tally() %>%
    arrange(desc(n)) %>%
    head(5)
```

Be careful not to include anything that would split the group into
subgroups, such as `sex`, `year` etc.

#### Challenge

1.  How many individuals were caught in each `plot_type` surveyed?

2.  You saw above how to count the number of individuals of each `sex`
    using a combination of `group_by()` and `tally()`. How could you get
    the same result using `group_by()` and `summarize()`? Hint: see
    `?n`.

```{r, include=FALSE}
## Answer 1
surveys %>%
    group_by(plot_type) %>%
    tally()

## Answer 2
surveys %>%
  group_by(sex) %>%
  summarize(n = n())
```

## Reshaping with pivot_wider and pivot_longer

### Defining wide vs long data

The survey data presented here is almost in what we call a *long* format
-- every observation of every individual is its own row. This is an
ideal format for data with a rich set of information per observation. It
makes it difficult, however, to look at the relationships between
measurements across plots/trials. For example, what is the relationship
between mean weights of different genera across all plots?

To answer that question, we want each plot to have its own row, with
each measurements in its own column. This is called a *wide* data
format. For the `surveys` data as we have it right now, this is going to
be one heck of a wide data frame! However, if we were to summarize data
within plots and species, we can reduce the dataset and begin to look
for some relationships we'd want to examine. We need to create a new
table where each row is the values for a particular variable associated
with each plot. In practical terms, this means the values in genus would
become the names of column variables and the cells would contain the
values of the mean weight observed on each plot by genus.

We can use the functions called `pivot_wider()` and `pivot_longer()`
(these are newer replacements for `spread()` and `gather()`, which were
the older functions). These can feel tricky to think through, but do not
feel alone in this! Many others have squinted at their data, unsure
exactly how to reshape it, so there are many
[guides](https://libguides.princeton.edu/R-reshape) and
[cheatsheets](https://bioinformatics.ccr.cancer.gov/docs/rintro/resources/tidyr_cheatsheet.pdf)
available to help!

### Summary of long vs wide formats

Long format:

-   every column is a variable
    -   first column(s) repeat
-   every row is an observation

Wide format:

-   each row is a measured thing
-   each column is an independent observation
    -   first column does not repeat

### Long to Wide with `pivot_wider`

Let's start by using `dplyr` to create a data frame with the mean body
weight of each genus by plot.

```{r}
surveys_gw <- surveys %>%
    filter(!is.na(weight)) %>%
    group_by(genus, plot_id) %>%
    summarize(mean_weight = mean(weight))

surveys_gw %>% head()
```

Now, to make this long data wide, we use `pivot_wider()` from `tidyr` to
spread out the different taxa into columns. `pivot_wider()` takes 3
arguments: the data , the `names_from` column variable that will
eventually become the column names, and the `values_from` column
variable that will fill in the values. We'll use a pipe so we don't need
to explicitly supply the data argument.

```{r}
surveys_gw_wide <- surveys_gw %>% 
  pivot_wider(names_from = genus, values_from = mean_weight)

head(surveys_gw_wide)
```

Now we can go back to our original question: what is the relationship
between mean weights of different genera across all plots? We can easily
see the weights for each genus in each plot! Notice that some genera
have `NA` values. That's because some genera were not recorded in that
plot.

Note! If you've been using R for a few years in other contexts, you may
have used the function `spread()`instead of pivot_wider. spread also
takes three arguments: the data, the *key* column (or column with
identifying information), and the *values* column (the one with the
numbers/values), and can return the same \`wide' data frame. However
pivot_wider is currently the recommended function to use, as it has
extra features to make it more flexible and less error_prone.

```{r}
surveys_gw_wide0 <- surveys_gw %>%
  spread(key = genus, value = mean_weight) 

head(surveys_gw_wide0)
```

### Wide to long with `gather` and `pivot_longer`

What if we had the opposite problem, and wanted to go from a wide to
long format? For that, we can use `pivot_longer()` to gather a set of
columns into one key-value pair. To go backwards from `surveys_gw_wide`,
we should exclude *plot_id*.

`pivot_longer()` takes 4 arguments: the data, the `names_to` column
variable that comes from the column names, the `values_to` column with
the values, and `cols` which specifies which columns we want to keep or
drop. Again, we will pipe from the dataset so we don't have to specify
the data argument:

```{r}
surveys_gw_long2 <- surveys_gw_wide %>% 
  pivot_longer(names_to = "genus", values_to = "mean_weight", cols = -plot_id)

surveys_gw_long2
```

If the columns are directly adjacent as they are here, we don’t even
need to list the all out: we can just use the : operator, as before.

```{r}
surveys_gw_wide %>% 
  pivot_longer(names_to = "genus", values_to = "mean_weight", cols = Baiomys:Sigmodon)
```

Note that now the `NA` genera are included in the long format.

In the past, you may have used `gather()`. We give it the arguments of a
new key and value column name, and then specify which columns we either
want or do not want gathered up. So, togo backwards from
`surveys_gw_wide`, and exclude `plot_id` from the gathering, we would do
the following:

```{r}
surveys_gw_long0 <- surveys_gw_wide0 %>%
  gather(genus, mean_weight, -plot_id) 

head(surveys_gw_long0)
```

### Challenge

Starting with the `surveys_gw_wide` dataset, how would you display a new
dataset that gathers the mean weight of all the genera (excluding NAs)
except for the genus `Perognathus`?

```{r}
surveys_gw_wide %>%
  pivot_longer(names_to = "genus", values_to = "mean_weight",
               cols = c(-plot_id, -Perognathus))
```

## 
