---
title: 'Introduction to R: Assignment, vectors, functions, strings, and loops'
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

*Note: Components of this lecture (and some others in the course!) were
originally created by voluntary contributions to [Data
Carpentry](https://datacarpentry.org) and have been modified and
expanded over the years to align with the aims of EEB313. Data Carpentry
is an organization focused on data literacy, with the objective of
teaching skills to researchers to enable them to retrieve, view,
manipulate, analyze, and store their and other’s data in an open and
reproducible way in order to extract knowledge from data.*

The above paragraph is made explicit since it is one of the core
features of working with an open language like R. Many smart people
willingly and actively share their material publicly, so that others can
modify and build off of the material themselves. See the [About
Us](about-us) page for more information on all the contributors to EEB
313.

------------------------------------------------------------------------

## Lesson Preamble

> ### Learning Objectives
>
> -   Define the following terms as they relate to R: call, function,
>     arguments, options.
> -   Use comments within code blocks.
> -   Do simple arithmetic operations in R using values and objects.
> -   Call functions and use arguments to change their default options.
> -   Define our own functions
> -   Inspect the content of vectors and manipulate their content.
> -   Create for-loops
> -   Describe what a data frame is.
> -   Load external data from a .csv file into a data frame in R.

------------------------------------------------------------------------

## Getting started in RStudio

Before starting this lecture, make sure you've completed [all the
steps](lectures/lec01-downloading-r) to download and install R and
RStudio, and get familiar with the layout of RStudio, including two of
the ways we can enter commands: in the console, and in an R script file.

For this lecture, we'll start off just entering commands at the
**console**, and then move on to writing longer multi-line code in an
**R script file**.

For later lectures and assignments, we'll go over and start to use the R
Markdown notebook format. These lecture notes are actually made using
one of these notebooks, so whenever we want to show some code, it will
be in a grey box - and we actually executed that code to make these
notes, so we know all our examples work as expected!

## Creating objects in R

As we saw in our first class, you can get output from R simply by typing
math in the console:

```{r}
3 + 5
```

```{r}
12 / 7
```

Most arithmetic operations can be done in R using intuitive notation,
and the regular rules of order-of-operations and use of brackets apply,
e.g.

```{r}
(1+2)^2 + (6/3)
```

However, to do useful and interesting things, we need to assign *values*
to *objects*.

```{r}
x <- 3
```

This command creates a new variable (a type of object) in R's memory,
and assigns it a value of 3. (The symbol `<-` does this assignment. You
can also use the `=` sign that other programming langauges use, but it
can sometimes have slightly different behaviour, so R users almost
always recommend using the `<-` symbol instead). If you look in the
`Environment` panel in the top right of RStudio, you should see `x`
listed along with its value.

We can now perform calculations using `x`; the symbol will be replaced
by the assigned value in every calculation.

```{r}
x + 5
```

You can name an object in R almost anything you want:

```{r}
joel <- 3
joel + 5
```

#### Challenge

So far, we have created two variables, `joel` and `x`. What is the sum
of these variables?

```{r, include=FALSE}
joel + x
```

### Some tips on naming objects

-   Objects can be given any name: `x`, `current_temperature`, `thing`,
    or `subject_id`.
-   You want your object names to be explicit and not too long.
-   Object names cannot start with a number: `x2` is valid, but `2x` is
    not.
-   R is also case sensitive: `joel` is different from `Joel`.
-   Avoid using the names of existing functions (e.g. `mean`, `df`). You
    can check whether the name is already in use by using tab completion
-   Generally good to use underscores (`_`) to separate words in
    variable and function names

It is also recommended to use nouns for variable names, and verbs for
function names. It's important to be consistent in the styling of your
code (where you put spaces, how you name variables, etc.). Using a
consistent coding style[^1] makes your code clearer to read for your
future self and your collaborators. RStudio will format code for you if
you highlight a section of code and press <kbd>Ctrl/Cmd</kbd> +
<kbd>Shift</kbd> + <kbd>a</kbd>.

[^1]: Refer to the [tidy style
    guide](https://style.tidyverse.org/index.html) for which style to
    adhere to.

### Preforming calculations

When assigning a value to an object, R does not print anything. You can
force R to print the value by using parentheses or by typing the object
name:

```{r}
weight_kg <- 55    # doesn't print anything
(weight_kg <- 55)  # but putting parentheses around the call prints the value of `weight_kg`
weight_kg          # and so does typing the name of the object
```

The variable `weight_kg` is stored in the computer's memory where R can
access it, and we can start doing arithmetic with it efficiently. For
instance, we may want to convert this weight into pounds (weight in
pounds is 2.2 times the weight in kg):

```{r}
2.2 * weight_kg
```

We can also change a variable's value by assigning it a new one:

```{r}
weight_kg <- 57.5
2.2 * weight_kg
```

This means that assigning a value to one variable does not change the
values of other variables. For example, let's store the animal's weight
in pounds in a new variable, `weight_lb`:

```{r}
weight_lb <- 2.2 * weight_kg # Actually, 1 kg = 2.204623 lbs
```

and then change `weight_kg` to 100.

```{r}
weight_kg <- 100
```

#### Challenge

What do you think is the current content of the object `weight_lb`?
126.5 or 220?

```{r, eval=FALSE}
weight_lb
```

#### Challenge

What are the values after each statement in the following?

```{r, eval=FALSE}
mass <- 47.5
age  <- 122
mass <- mass * 2.0      # mass?
age  <- age - 20        # age?
mass_index <- mass/age  # mass_index?
```

## Functions and their arguments

### Understanding functions

Functions can be thought of as recipes. You give a few ingredients as
input to a function, and it will generate an output based on these
ingredients. Just as with baking, both the ingredients and the actual
recipe will influence what comes out of the recipe in the end: will it
be a cake or a loaf of bread? In R, the inputs to a function are not
called ingredients, but rather *arguments*, and the output is called the
*return value* of the function. A function does not technically have to
return a value, but often does so. Functions are used to automate more
complicated sets of commands and many of them are already predefined in
R. A typical example would be the function `sqrt()`. The input (the
argument) must be a number, and the return value (in fact, the output)
is the square root of that number. Executing a function ('running it')
is called *calling* the function. An example of a function call is:

```{r}
sqrt(9)
```

Which is the same as assigning the value to a variable and then passing
that variable to the function:

```{r}
a <- 9
b <- sqrt(a)
b
```

Here, the value of `a` is given to the `sqrt()` function, the `sqrt()`
function calculates the square root, and returns the value which is then
assigned to variable `b`. This function is very simple, because it takes
just one argument.

The return 'value' of a function need not be numerical (like that of
`sqrt()`), and it also does not need to be a single item: it can be a
set of things, or even a dataset, as we will see later on.

Arguments can be anything, not only numbers or filenames, but also other
objects. Exactly what each argument means differs per function, and must
be looked up in the documentation (see below). Some functions take
arguments which may either be specified by the user, or, if left out,
take on a *default* value: these are called *options*. Options are
typically used to alter the way the function operates, such as whether
it ignores 'bad values', or what symbol to use in a plot. However, if
you want something specific, you can specify a value of your choice
which will be used instead of the default.

### Tab-completion

To access help about `sqrt`, we are first going to learn about
tab-completion. Type `s` and press <kbd>Tab</kbd>.

```{r, eval=FALSE}
s<tab>q
```

You can see that R gives you suggestions of what functions and variables
are available that start with the letter `s`, and thanks to RStudio they
are formatted in this nice list. There are *many* suggestions here, so
let's be a bit more specific and append a `q`, to find what we want. If
we press enter or tab again, R will insert the selected option.

You can see that R inserts a pair of parentheses together with the name
of the function. This is how the function syntax looks for R and many
other programming languages, and it means that within these parentheses,
we will specify all the arguments (the ingredients) that we want to pass
to this function.

If we press tab again, R will helpfully display all the available
parameters for this function that we can pass an argument to. The word
*parameter* is used to describe the name that the argument can be passed
to. More on that later.

```{r, eval=FALSE}
sqrt(<tab>
```

There are many things in this list, but only one of them is marked in
purple. Purple here means that this list item is a parameter we can use
for the function, while yellow means that it is a variable that we
defined earlier.[^2]

[^2]: There are a few other symbols as well, all of which can be viewed
    at the end of [this post about RStudio code
    completion](https://support.rstudio.com/hc/en-us/articles/205273297-Code-Completion).

### Help with defined functions

To read the full help about `sqrt`, we can use the question mark, or
type it directly into the help document browser.

```{r, eval=FALSE}
?sqrt
```

As you can see, `sqrt()` takes only one argument, `x`, which needs to be
a *numerical vector*. Don't worry too much about the fact that it says
*vector* here; we will talk more about that later. Briefly, a numerical
vector is one or more numbers. In R, every number is a vector, so you
don't have to do anything special to create a vector. More on vectors
later.

Let's try a function that can take multiple arguments: `round()`.

```{r, eval=FALSE}
round(<tab>)
?round
```

If we try round with a value:

```{r}
round(3.14159)
```

Here, we've called `round()` with just one argument, `3.14159`, and it
has returned the value `3`. That's because the default is to round to
the nearest whole number, or integer. If we want more digits we can pass
an argument to the `digits` parameter, to specify how many decimals we
want to round to.

```{r}
round(3.14159, digits = 2)
```

So, above we pass the *argument* `2`, to the *parameter* `digits`.
Knowing this nomenclature is not essential for doing your own data
analysis, but it will be very helpful when you are reading through help
documents online and in RStudio.

We can leave out the word `digits` since we know it comes as the second
parameter, after `x`.

```{r}
round(3.14159, 2)
```

As you notice, we have been leaving out `x` from the beginning. If you
provide the names for both the arguments, we can switch their order:

```{r}
round(digits = 2, x = 3.14159)
```

It's good practice to put the non-optional arguments (like the number
you're rounding) first in your function call, and to specify the names
of all optional arguments. If you don't, someone reading your code might
have to look up the definition of a function with unfamiliar arguments
to understand what you're doing.

## Writing functions

In this class, you will be working a lot with functions, especially
those that someone else has already written. When you type `sum`, `c()`,
or `mean()`, you are using a function that has been made previously and
built into R. To remove some of the magic around these functions, we
will go through how to make a basic function of our own. Let's start
with a simple example where we add two numbers together:

```{r}
add_two_numbers <- function(num1, num2) {
    return(num1 + num2)
}
add_two_numbers(4, 5)
```

As you can see, running this function on two numbers returns their sum.
We could also assign to a variable in the function and return the
function.

```{r}
add_two_numbers <- function(num1, num2) {
    my_sum <- num1 + num2
    return(my_sum)
}
add_two_numbers(4, 5)
```

\#### Challenge

Can you write a function that calculates the mean of 3 numbers?

```{r, include=FALSE}
mean_of_three_numbers <- function(num1, num2, num3) {
   my_sum <- num1 + num2 + num3
   my_mean <- my_sum / 3
   return(my_mean)
}
mean_of_three_numbers(2, 4, 6)
```

## Vectors and data types

A vector is the most common and basic data type in R, and is pretty much
the workhorse of R. A vector is composed by a series of values, which
can be either numbers or characters. We can assign a series of values to
a vector using the `c()` function, which stands for "concatenate
(combine/connect one after another) values into a vector" For example we
can create a vector of animal weights and assign it to a new object
`weight_g`:

```{r}
weight_g <- c(50, 60, 65, 82) # Concatenate/Combine values into a vector
weight_g
```

You can also use the built-in command `seq`, to create a sequence of
numbers without typing all of them in manually.

```{r}
seq(0, 30) # This is the same as just `0:30`
seq(0, 30, 3) # Every third number
```

A vector can also contain characters:

```{r}
animals <- c('mouse', 'rat', 'dog')
animals
```

The quotes around "mouse", "rat", etc. are essential here and can be
either single or double quotes. Without the quotes R will assume there
are objects called `mouse`, `rat` and `dog`. As these objects don't
exist in R's memory, there will be an error message.

There are many functions that allow you to inspect the content of a
vector. `length()` tells you how many elements are in a particular
vector:

```{r}
length(weight_g)
length(animals)
```

An important feature of a vector is that all of the elements are the
same type of data. The function `class()` indicates the class (the type
of element) of an object:

```{r}
class(weight_g)
class(animals)
```

The function `str()` provides an overview of the **str**ucture of an
object and its elements. It is a useful function when working with large
and complex objects:

```{r}
str(weight_g)
str(animals)
```

You can use the `c()` function to add other elements to your vector:

```{r}
weight_g <- c(weight_g, 90) # add to the end of the vector
weight_g <- c(30, weight_g) # add to the beginning of the vector
weight_g
```

In the first line, we take the original vector `weight_g`, add the value
`90` to the end of it, and save the result back into `weight_g`. Then we
add the value `30` to the beginning, again saving the result back into
`weight_g`.

We can do this over and over again to grow a vector, or assemble a
dataset. As we program, this may be useful to add results that we are
collecting or calculating.

An **atomic vector** is the simplest R **data type** and it is a linear
vector of a single type, e.g. all numbers. Above, we saw 2 of the 6 main
**atomic vector** types that R uses: `"character"` and `"numeric"` (or
`"double"`). These are the basic building blocks that all R objects are
built from.

Vectors are one of the many **data structures** that R uses. Other
important ones are lists (`list`), matrices (`matrix`), data frames
(`data.frame`), factors (`factor`) and arrays (`array`). In this class,
we will focus on data frames, which is most commonly used one for data
analyses.

#### Challenge

We’ve seen that atomic vectors can be of type character, numeric (or
double), integer, and logical. But what happens if we try to mix these
types in a single vector? Find out by using `class` to test these
examples.

```{r}
num_char <- c(1, 2, 3, 'a')
num_logical <- c(1, 2, 3, TRUE)
char_logical <- c('a', 'b', 'c', TRUE)
tricky <- c(1, 2, 3, '4')
```

```{r, include=FALSE}
# Answer
class(num_char)
class(num_logical)
class(char_logical)
class(tricky)
```

This happens because vectors can be of only one data type. Instead of
throwing an error and saying that you are trying to mix different types
in the same vector, R tries to convert (coerce) the content of this
vector to find a "common denominator". A logical can be turn into 1 or
0, and a number can be turned into a string/character representation. It
would be difficult to do it the other way around: would 5 be TRUE or
FALSE? What number would 't' be?

In R, we call converting objects from one class into another class
*coercion*. These conversions happen according to a hierarchy, whereby
some types get preferentially coerced into other types. Can you draw a
diagram that represents the hierarchy of how these data types are
coerced?

```{r, include=FALSE}
# Answer:
# logical -> numeric -> character <- logical
```

This can be important to watch for in data sets that you import.

## Subsetting vectors

If we want to extract one or several values from a vector, we must
provide one or several indices in square brackets. For instance:

```{r}
animals <- c("mouse", "rat", "dog", "cat")
animals[2]
animals[c(3, 2)]
```

We can also repeat the indices to create an object with more elements
than the original one:

```{r}
more_animals <- animals[c(1, 2, 3, 2, 1, 4)]
more_animals
```

R indices start at 1. Programming languages like Fortran, MATLAB, Julia,
and R start counting at 1, because that's what human beings typically
do. Languages in the C family (including C++, Java, Perl, and Python)
count from 0 because that was historically simpler for computers and can
allow for more elegant code.

### Conditional subsetting

Another common way of subsetting is by using a logical vector. `TRUE`
will select the element with the same index, while `FALSE` will not:

```{r}
weight_g <- c(21, 34, 39, 54, 55)
weight_g[c(TRUE, FALSE, TRUE, TRUE, FALSE)]
```

Typically, these logical vectors are not typed by hand, but are the
output of other functions or logical tests. For instance, if you wanted
to select only the values above 50:

```{r}
weight_g > 50    # will return logicals with TRUE for the indices that meet the condition
## so we can use this to select only the values above 50
weight_g[weight_g > 50]
```

We will consider conditions in more detail in the next few lectures.

### Strings (character vectors)

Just a small note about character vectors, also called `strings`. There
are built-in packages for subsetting them that we'll learn about later.
They can be particularly relevant for ecological/genomics because
important data can be nested in complicated strings of text (ex:
extracting only the observations that occurred in wet habitats from a
column of habitat descriptions or only genes with functions related to
drought tolerance).

```{r}
string1 <- "This is a string" # you can include spaces between your quotes
string2 <- c(string1, "so is this") # concatenate with another string
string2[2] # can access the second string via subsetting

# Playing a bit with declaring variables
"You can include 'quotes' in a string"
string3 <- 'You can include "quotes" in a string'
string3
"You can include \"matching quotes\" if you 'escape' them with a backslash (\\)"
```

## Missing data

As R was designed to analyze datasets, it includes the concept of
missing data (which is uncommon in other programming languages). Missing
data are represented in vectors as `NA`.

When doing operations on numbers, most functions will return `NA` if the
data you are working with include missing values. This feature makes it
harder to overlook the cases where you are dealing with missing data.
You can add the argument `na.rm = TRUE` to calculate the result while
ignoring the missing values.

```{r}
heights <- c(2, 4, 4, NA, 6)
mean(heights)
max(heights)
mean(heights, na.rm = TRUE)
max(heights, na.rm = TRUE)
```

```{r}
## Extract those elements which are not missing values.
heights[!is.na(heights)]

## Returns the object with incomplete cases removed. The returned object is an atomic vector of type `"numeric"` (or `"double"`).
na.omit(heights)

## Extract those elements which are complete cases. The returned object is an atomic vector of type `"numeric"` (or `"double"`).
heights[complete.cases(heights)]
```

Recall that you can use the `class()` function to find the type of your
atomic vector.

#### Challenge

1.  Using this vector of length measurements, create a new vector with
    the NAs removed.

```{r}
lengths <- c(10, 24, NA, 18, NA, 20)
```

```{r, include=FALSE}
# Answer
na.omit(lengths)
```

2.  Use the function `median()` to calculate the median of the `lengths`
    vector.

```{r, include=FALSE}
# Answer
median(lengths, na.rm = TRUE)
```

## Loops and vectorization

Loops, specifically for-loops, are essential to programming in general.
However, in R, you should avoid them as often as possible because there
are more efficient ways of doing things that you should use instead. It
is still important that you understand the concept of loops and you
might also use them in some of your own functions if there is no
vectorized way of going about what you want to do.

You can think of a for-loop as: "for each number contained in a
list/vector, perform this operation" and the syntax basically says the
same thing:

```{r}
v <- c(2, 4, 6)
for (num in v) {
    print(num)
}
```

Instead of printing out every number to the console, we could also add
numbers cumulatively, to calculate the sum of all the numbers in the
vector:

```{r}
# To increment `w` each time, we must first create the variable,
# which we do by setting `w <- 0`, referred to as initializing.
# This also ensures that `w` is zero at the start of the loop and
# doesn't retain the value from last time we ran this code.
w <- 0
for (num in v) {
    w <- w + num
}
w
```

If we put what we just did inside a function, we have essentially
recreated the `sum` function in R.

```{r}
my_sum <- function(input_vector) {
    vector_sum <- 0
    for (num in input_vector){
        vector_sum <- vector_sum + num
    }
    return(vector_sum)
}

my_sum(v)
```

Although this gives us the same output as the built-in function `sum`,
the built-in function has many more optimizations so it is much faster
than our function. In R, it is always faster to try to find a way of
doing things without writing a loop yourself. When you are reading about
R, you might see suggestions that you should try to *vectorize* your
code to make it faster. What people are referring to, is that you should
not write for loops in R and instead use the ready-made functions that
are much more efficient in working with vectors and essentially performs
operations on entire vector at once instead of one number at a time.
Conceptually, loops operate on one element at a time while vectorized
code operates on all elements of a vector at once.
